---
title: "Classification Analysis: Airline Passenger Satisfaction"
author: "DSC532 Group Project"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    theme: cosmo
    code_folding: show
    fig_width: 10
    fig_height: 6
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, warning = FALSE, message = FALSE,
  fig.align = "center", fig.width = 10, fig.height = 6
)
```

# Setup and Libraries

```{r libraries}
# Install packages if needed (uncomment on first run)
# install.packages(c("tidyverse", "janitor", "skimr", "corrplot", "caret",
#                     "MASS", "class", "e1071", "glmnet", "pROC",
#                     "knitr", "kableExtra", "gridExtra", "boot"))

library(tidyverse)
library(janitor)
library(skimr)
library(corrplot)
library(caret)
library(MASS)
library(class)
library(e1071)
library(glmnet)
library(pROC)
library(knitr)
library(kableExtra)
library(gridExtra)
library(boot)

set.seed(42)
```
```{r}
getwd()
```
# Data Loading and Overview

The dataset contains airline passenger satisfaction survey responses. We load the pre-split training and test sets separately to preserve the integrity of the evaluation pipeline — the test set will not be touched until the final model evaluation.

```{r load-data}
train_raw <- read_csv("train.csv")
test_raw  <- read_csv("test.csv")

cat("Training set dimensions:", dim(train_raw), "\n")
cat("Test set dimensions:", dim(test_raw), "\n")
```
```{r first-look}
glimpse(train_raw)
```

```{r head-data}
head(train_raw)
```

```{r summary-data}
summary(train_raw)
```

The training set has `r nrow(train_raw)` observations and `r ncol(train_raw)` variables. The first two columns (`...1` and `id`) are administrative identifiers with no predictive value. The target variable `satisfaction` is binary: "satisfied" or "neutral or dissatisfied".

# Data Cleaning

## Removing Identifiers and Standardising Column Names

We remove the unnamed index column (`...1`) and the `id` column since they carry no predictive information, and apply `clean_names()` to ensure consistent snake_case naming throughout the analysis.

```{r clean-names}
train_clean <- train_raw %>%
  clean_names() %>%
  dplyr::select(-any_of(c("x1", "x_1")), -id)

test_clean <- test_raw %>%
  clean_names() %>%
  dplyr::select(-any_of(c("x1", "x_1")), -id)

colnames(train_clean)
```

After cleaning we have `r ncol(train_clean)` variables: 4 continuous (age, flight distance, departure delay, arrival delay), 14 ordinal service ratings (1–5 scale), and 4 categorical variables (gender, customer type, type of travel, class), plus the target.

## Missing Values Analysis

```{r missing-values}
bind_rows(
  train_clean %>%
    summarise(across(everything(), ~ sum(is.na(.)))) %>%
    pivot_longer(everything(), names_to = "variable", values_to = "na_count") %>%
    mutate(dataset = "train", na_percent = round(na_count / nrow(train_clean) * 100, 3)),

  test_clean %>%
    summarise(across(everything(), ~ sum(is.na(.)))) %>%
    pivot_longer(everything(), names_to = "variable", values_to = "na_count") %>%
    mutate(dataset = "test", na_percent = round(na_count / nrow(test_clean) * 100, 3))
) %>%
  filter(na_count > 0) %>%
  dplyr::select(dataset, variable, na_count, na_percent) %>%
  arrange(desc(na_count)) %>%
  kable(col.names = c("Dataset", "Variable", "Missing Count", "Missing %")) %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

Missing values are present exclusively in `arrival_delay_in_minutes` — roughly 0.3% of both the training and test sets. This low proportion rules out systematic dropout but we still need a principled imputation strategy rather than simple deletion. Before deciding on an approach, we conduct a thorough EDA to understand the structure of the data and the relationships between all variables — in particular with `arrival_delay_in_minutes`.

# Exploratory Data Analysis

The goal of EDA is threefold: (1) understand the distributions of individual variables and their relationship with the target, (2) identify any data quality issues, and (3) gather the evidence needed to justify our imputation strategy for the missing arrival delay values.

## Target Variable Distribution

```{r target-dist}
train_clean %>%
  count(satisfaction) %>%
  mutate(prop = round(n / sum(n) * 100, 1)) %>%
  kable(col.names = c("Class", "Count", "Percentage (%)")) %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

```{r target-barplot, fig.height=4}
train_clean %>%
  count(satisfaction) %>%
  mutate(percent = n / sum(n),
         label = scales::percent(percent)) %>%
  ggplot(aes(x = "", y = n, fill = satisfaction)) +
  geom_col(width = 1) +
  coord_polar(theta = "y") +
  geom_text(aes(label = label), position = position_stack(vjust = 0.5), size = 5) +
  labs(title = "Passenger Satisfaction Distribution") +
  theme_void()
```

The dataset exhibits a moderate class imbalance: approximately 56.7% of passengers are neutral or dissatisfied and 43.3% are satisfied. This imbalance is not severe enough to require resampling techniques, but it means raw accuracy can be misleading — we will prioritise F1 score, precision, recall, and AUC when comparing models.

## Distributions of Continuous Variables

```{r cont-dist-overall, fig.height=5}
train_clean %>%
  dplyr::select(age, flight_distance,
                departure_delay_in_minutes, arrival_delay_in_minutes) %>%
  pivot_longer(everything(), names_to = "variable", values_to = "value") %>%
  ggplot(aes(x = value)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "white", alpha = 0.8) +
  facet_wrap(~ variable, scales = "free", ncol = 2) +
  labs(title = "Distributions of Continuous Variables",
       x = "Value", y = "Count") +
  theme_minimal()
```

**Age** is moderately spread across adults, with a slight bimodal pattern suggesting two main passenger age groups.

**Flight distance** is right-skewed, with the majority of flights being short-to-medium haul and a long upper tail likely for intercontinental routes.

**Departure and arrival delay** are both heavily right-skewed with a mass of near-zero values and a sparse but long upper tail. These will be examined in detail in the outlier analysis below.

```{r cont-dist-by-sat, fig.height=5}
train_clean %>%
  dplyr::select(satisfaction, age, flight_distance,
                departure_delay_in_minutes, arrival_delay_in_minutes) %>%
  pivot_longer(cols = -satisfaction, names_to = "variable", values_to = "value") %>%
  ggplot(aes(x = value, fill = satisfaction)) +
  geom_histogram(bins = 40, color = "white", alpha = 0.6, position = "identity") +
  facet_wrap(~ variable, scales = "free", ncol = 2) +
  labs(title = "Continuous Variable Distributions by Satisfaction Class") +
  theme_minimal()
```

Breaking the distributions down by satisfaction class reveals two notable patterns. First, satisfied passengers tend to be somewhat older, with the satisfied distribution shifted towards middle age. Second, satisfied passengers are more heavily concentrated among longer flights, suggesting that business travellers (who tend to fly further) drive satisfaction. Departure and arrival delays do not show a strong visual separation by class, which is consistent with the finding that most passengers experience little to no delay.

### Delay Distribution Analysis

```{r delay-zoom}
p_dep <- train_clean %>%
  ggplot(aes(x = departure_delay_in_minutes, fill = satisfaction)) +
  geom_histogram(binwidth = 5, alpha = 0.6, position = "identity") +
  coord_cartesian(xlim = c(0, 200)) +
  labs(title = "Departure Delay (0–200 min)", x = "Minutes", y = "Count") +
  theme_minimal()

p_arr <- train_clean %>%
  ggplot(aes(x = arrival_delay_in_minutes, fill = satisfaction)) +
  geom_histogram(binwidth = 5, alpha = 0.6, position = "identity") +
  coord_cartesian(xlim = c(0, 200)) +
  labs(title = "Arrival Delay (0–200 min)", x = "Minutes", y = "Count") +
  theme_minimal()

grid.arrange(p_dep, p_arr, ncol = 2)
```

Both delay variables have extreme right tails. We investigate whether the very high values are genuine extreme events or data entry errors.

### Delay Outlier Investigation

```{r delay-tail-analysis}
train_clean %>%
  dplyr::select(departure_delay_in_minutes, arrival_delay_in_minutes) %>%
  summarise(across(everything(), list(
    min    = ~ min(., na.rm = TRUE),
    p25    = ~ quantile(., 0.25, na.rm = TRUE),
    median = ~ median(., na.rm = TRUE),
    p75    = ~ quantile(., 0.75, na.rm = TRUE),
    p95    = ~ quantile(., 0.95, na.rm = TRUE),
    p99    = ~ quantile(., 0.99, na.rm = TRUE),
    max    = ~ max(., na.rm = TRUE)
  ))) %>%
  pivot_longer(everything(), names_to = c("variable", "stat"),
               names_sep = "_(?=[^_]+$)") %>%
  pivot_wider(names_from = stat, values_from = value) %>%
  kable(digits = 1) %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

```{r extreme-counts}
train_clean %>%
  summarise(
    dep_over_180  = sum(departure_delay_in_minutes > 180, na.rm = TRUE),
    dep_over_300  = sum(departure_delay_in_minutes > 300, na.rm = TRUE),
    dep_over_500  = sum(departure_delay_in_minutes > 500, na.rm = TRUE),
    arr_over_180  = sum(arrival_delay_in_minutes > 180, na.rm = TRUE),
    arr_over_300  = sum(arrival_delay_in_minutes > 300, na.rm = TRUE),
    arr_over_500  = sum(arrival_delay_in_minutes > 500, na.rm = TRUE)
  ) %>%
  kable() %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

```{r extreme-rows}
train_clean %>%
  filter(departure_delay_in_minutes > 500) %>%
  dplyr::select(departure_delay_in_minutes, arrival_delay_in_minutes,
                flight_distance, class, type_of_travel,
                departure_arrival_time_convenient) %>%
  arrange(desc(departure_delay_in_minutes))
```

Inspecting the extreme cases confirms that very long departure delays are always accompanied by comparably long arrival delays — the two variables move in lockstep. Delays exceeding 500 minutes (over 8 hours) are operationally plausible and can occur in practice due to severe weather, mechanical failures, or flight rescheduling to the following day. We therefore retain these observations, as removing them would introduce selection bias and prevent our models from generalising to such rare but real scenarios.

## Flight Distance Outlier Analysis

```{r flight-dist-quantiles}
train_clean %>%
  dplyr::select(flight_distance) %>%
  summarise(across(everything(), list(
    min    = ~ min(., na.rm = TRUE),
    p25    = ~ quantile(., 0.25, na.rm = TRUE),
    median = ~ median(., na.rm = TRUE),
    p75    = ~ quantile(., 0.75, na.rm = TRUE),
    p95    = ~ quantile(., 0.95, na.rm = TRUE),
    p99    = ~ quantile(., 0.99, na.rm = TRUE),
    max    = ~ max(., na.rm = TRUE)
  ))) %>%
  pivot_longer(everything(),
               names_to = c("variable", "stat"),
               names_sep = "_(?=[^_]+$)") %>%
  pivot_wider(names_from = stat, values_from = value) %>%
  kable(digits = 1) %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

While the upper tail of `flight_distance` reaches 4,983 miles, this is entirely consistent with long-haul intercontinental routes. No implausible values were identified and no observations are removed.

## Service Rating Distributions

The 14 service attributes are measured on a discrete 1–5 scale (with 0 occasionally representing "not applicable").

```{r rating-dist, fig.height=9}
rating_vars <- c("inflight_wifi_service", "departure_arrival_time_convenient",
                  "ease_of_online_booking", "gate_location", "food_and_drink",
                  "online_boarding", "seat_comfort", "inflight_entertainment",
                  "on_board_service", "leg_room_service", "baggage_handling",
                  "checkin_service", "inflight_service", "cleanliness")

train_clean %>%
  dplyr::select(satisfaction, all_of(rating_vars)) %>%
  pivot_longer(cols = -satisfaction, names_to = "variable", values_to = "value") %>%
  ggplot(aes(x = factor(value), fill = satisfaction)) +
  geom_bar(position = "stack") +
  facet_wrap(~ variable, scales = "free_y", ncol = 4) +
  labs(title = "Service Rating Distributions by Satisfaction",
       x = "Rating (1–5)", y = "Count") +
  theme_minimal() +
  theme(strip.text = element_text(size = 7))
```

Most service attributes concentrate in the 3–4 range, reflecting a broadly moderate to positive perception of airline services. Notably, variables such as `inflight_entertainment`, `seat_comfort`, and `online_boarding` show a clearer split between the two satisfaction classes at the higher rating values — passengers rating these services as 4 or 5 are predominantly satisfied. Variables like `gate_location` and `departure_arrival_time_convenient` show less separation, suggesting they are weaker predictors of overall satisfaction.

## Satisfaction by Key Categorical Variables

```{r satisfaction-by-cat, fig.height=7}
p1 <- train_clean %>%
  count(class, satisfaction) %>%
  group_by(class) %>%
  mutate(prop = n / sum(n)) %>%
  ggplot(aes(x = class, y = prop, fill = satisfaction)) +
  geom_col(position = "dodge") +
  labs(title = "Satisfaction by Travel Class", y = "Proportion", x = "") +
  theme_minimal()

p2 <- train_clean %>%
  count(type_of_travel, satisfaction) %>%
  group_by(type_of_travel) %>%
  mutate(prop = n / sum(n)) %>%
  ggplot(aes(x = type_of_travel, y = prop, fill = satisfaction)) +
  geom_col(position = "dodge") +
  labs(title = "Satisfaction by Travel Type", y = "Proportion", x = "") +
  theme_minimal()

p3 <- train_clean %>%
  count(customer_type, satisfaction) %>%
  group_by(customer_type) %>%
  mutate(prop = n / sum(n)) %>%
  ggplot(aes(x = customer_type, y = prop, fill = satisfaction)) +
  geom_col(position = "dodge") +
  labs(title = "Satisfaction by Customer Type", y = "Proportion", x = "") +
  theme_minimal()

p4 <- train_clean %>%
  count(gender, satisfaction) %>%
  group_by(gender) %>%
  mutate(prop = n / sum(n)) %>%
  ggplot(aes(x = gender, y = prop, fill = satisfaction)) +
  geom_col(position = "dodge") +
  labs(title = "Satisfaction by Gender", y = "Proportion", x = "") +
  theme_minimal()

grid.arrange(p1, p2, p3, p4, ncol = 2)
```

The categorical variables reveal several strong patterns. Business class passengers are the most satisfied group by a wide margin, while Economy and Eco Plus passengers are predominantly dissatisfied — suggesting that cabin class is one of the strongest predictors in the dataset. Similarly, passengers travelling for business purposes show much higher satisfaction than those travelling for personal reasons. Loyal customers are more satisfied than disloyal customers, possibly because loyalty status correlates with business travel and premium cabin access. Gender shows only a negligible difference in satisfaction rates.

## Correlation Analysis

### Correlation Heatmap

```{r corr-numeric}
train_clean_num <- train_clean %>%
  mutate(satisfaction_num = ifelse(satisfaction == "satisfied", 1, 0)) %>%
  dplyr::select(where(is.numeric))

cor_mat <- cor(train_clean_num, use = "pairwise.complete.obs")
round(cor_mat, 2)
```

```{r corr-heatmap-numeric, fig.height=9, fig.width=10}
corrplot(cor_mat,
         method = "color",
         order = "hclust",
         number.cex = 0.55,
         tl.cex = 0.75,
         title = "Correlation Matrix of Numeric Variables",
         mar = c(0, 0, 1.5, 0))
```

A few key observations from the correlation matrix:

- **Departure delay and arrival delay** are almost perfectly correlated (r ≈ 0.97), confirming that the two variables are nearly interchangeable from a linear standpoint. This is the foundation for our imputation strategy.
- **`satisfaction_num`** (the binary-encoded target, 1 = satisfied, 0 = dissatisfied) shows moderate positive correlations with several service rating variables — particularly `online_boarding`, `inflight_entertainment`, and `seat_comfort` — and a moderate positive correlation with `flight_distance`. It is essentially uncorrelated with the delay variables, suggesting that delays alone are not the primary driver of dissatisfaction in this dataset.
- The service rating variables form a loose cluster of moderate positive inter-correlations (r ≈ 0.2–0.5), which is expected: passengers who rate one service dimension highly tend to rate others highly too. However, these correlations are not high enough to cause serious multicollinearity concerns.

Note: interpreting the correlation of binary variables (like `satisfaction_num`) with continuous predictors should be done with caution, as Pearson correlation assumes linearity and does not capture threshold or interaction effects.


## Arrival Delay: Justifying the Imputation Strategy

Before imputing the ~0.3% of missing `arrival_delay_in_minutes` values, we systematically examine its relationship with every other variable. The goal is to determine the minimal sufficient predictor set for the imputation model.

### Relationship with Numeric Variables

```{r arrival-corr-bar, fig.height=4}
arr_cors <- cor_mat["arrival_delay_in_minutes", ]
arr_cors_df <- data.frame(
  variable = names(arr_cors),
  correlation = arr_cors
) %>%
  filter(variable != "arrival_delay_in_minutes") %>%
  arrange(abs(correlation))

ggplot(arr_cors_df, aes(x = reorder(variable, abs(correlation)),
                        y = correlation, fill = abs(correlation) > 0.1)) +
  geom_col() +
  coord_flip() +
  scale_fill_manual(values = c("FALSE" = "steelblue", "TRUE" = "tomato"),
                    labels = c("Near-zero", "Meaningful")) +
  labs(title = "Correlation of Arrival Delay with All Other Numeric Variables",
       x = "", y = "Pearson r", fill = "|r| > 0.1") +
  theme_minimal()
```

The bar chart makes the pattern unmistakably clear: `departure_delay_in_minutes` has a near-perfect correlation with `arrival_delay_in_minutes` (r ≈ 0.97), while every other numeric variable — including age, flight distance, and all 14 service ratings — has an absolute correlation effectively equal to zero (|r| < 0.05). This means that once we account for departure delay, no additional numeric variable provides meaningful predictive power for the missing arrival delay values.

### Relationship with Categorical Variables

```{r arrival-delay-boxplots-zoomed}
train_clean %>%
  dplyr::select(arrival_delay_in_minutes, satisfaction, gender,
                customer_type, type_of_travel, class) %>%
  pivot_longer(-arrival_delay_in_minutes,
               names_to = "variable", values_to = "category") %>%
  ggplot(aes(x = category, y = arrival_delay_in_minutes, fill = category)) +
  geom_boxplot(outlier.alpha = 0.1) +
  facet_wrap(~ variable, scales = "free_x") +
  coord_cartesian(ylim = c(0, 180)) +
  theme_minimal() +
  theme(legend.position = "none") +
  labs(title = "Arrival Delay by Categorical Variables (Zoomed to 0–180 min)",
       y = "Arrival Delay (minutes)", x = "")
```


### Imputation Justification Summary

Combining the numerical and categorical evidence:

1. **Departure delay** is the only variable with a meaningful relationship with arrival delay (Pearson r ≈ 0.97).
2. All other **numeric variables** have near-zero correlations (|r| < 0.05).
3. **Categorical variables** show no practically meaningful differences in arrival delay across their levels.

Therefore, a simple linear regression `arrival_delay ~ departure_delay` is the most appropriate and parsimonious imputation model. Adding further predictors would contribute negligible information while increasing model complexity unnecessarily.

# Missing Value Imputation

## Imputation Model: Simple Linear Regression

Based on the EDA evidence, we fit a simple linear regression of arrival delay on departure delay using all complete cases in the **training set only**. Fitting on the training set prevents data leakage — the imputation model sees no information from the test set.

```{r imputation-model}
train_reg <- train_clean %>%
  dplyr::select(departure_delay_in_minutes, arrival_delay_in_minutes) %>%
  na.omit()

# Fit the imputation model on complete training cases
delay_model <- lm(arrival_delay_in_minutes ~ departure_delay_in_minutes,
                  data = train_reg)
summary(delay_model)
```

The model summary reveals an extremely strong fit: the R² is approximately 0.94, meaning that departure delay alone explains 94% of the variance in arrival delay. The F-statistic is enormous and the p-value for the coefficient is effectively zero, confirming that the linear relationship is both highly significant and practically strong. The regression coefficient (~1.02) indicates that each additional minute of departure delay translates into approximately one additional minute of arrival delay on average — an intuitive and operationally sensible result.

```{r imputation-scatter, fig.height=4}
ggplot(train_reg, aes(x = departure_delay_in_minutes, y = arrival_delay_in_minutes)) +
  geom_point(alpha = 0.2, color = "steelblue", size = 0.5) +
  geom_smooth(method = "lm", color = "red", se = TRUE) +
  coord_cartesian(xlim = c(0, 200), ylim = c(0, 200)) +
  labs(title = "Arrival Delay vs Departure Delay with Fitted Regression Line",
       subtitle = paste("Pearson r =",
                        round(cor(train_reg$departure_delay_in_minutes,
                                  train_reg$arrival_delay_in_minutes), 4)),
       x = "Departure Delay (min)", y = "Arrival Delay (min)") +
  theme_minimal()
```

## Cross-Validation of Imputation Accuracy

To get an honest estimate of how well the imputation model generalises (rather than relying on in-sample R²), we perform 5-fold cross-validation and report the mean squared error.

```{r imputation-cv}
delay_model_glm <- glm(arrival_delay_in_minutes ~ departure_delay_in_minutes,
                       data = train_reg)
set.seed(123)
cv_result <- cv.glm(train_reg, delay_model_glm, K = 10)

cat("10-fold CV Mean Squared Error (imputation model):", round(cv_result$delta[1], 4), "\n")
cat("Root MSE:", round(sqrt(cv_result$delta[1]), 2), "minutes\n")
```

The cross-validated RMSE gives us a realistic estimate of the prediction error when imputing a missing arrival delay value from the corresponding departure delay. Given that departure delays span from 0 to over 1,500 minutes, an RMSE on the order of a few minutes confirms that the imputation model is highly accurate and will introduce only negligible noise into the imputed values.

## Applying the Imputation

```{r keep-raw-versions}
# Preserve the versions with missing values for reference
train_with_na <- train_clean
test_with_na  <- test_clean
```

```{r apply-imputation}
# Impute using the model fitted on training data only (no leakage)
train_clean$arrival_delay_in_minutes[is.na(train_clean$arrival_delay_in_minutes)] <-
  predict(delay_model,
          newdata = train_clean[is.na(train_clean$arrival_delay_in_minutes), ])

cat("Remaining NAs in train:", sum(is.na(train_clean$arrival_delay_in_minutes)), "\n")
```

After imputation the training set is complete. We impute only the training data here for exploratory analysis; the test set (`test_with_na`) is left untouched and will be preprocessed within the model evaluation pipeline to avoid any data leakage.

# Classification Modelling

We now proceed to classification. Our response variable is `satisfaction` (binary: "satisfied" vs "neutral or dissatisfied"). R modelling functions such as `glm()`, `lda()`, and `qda()` automatically handle categorical predictors by creating the appropriate dummy variables internally. For `class` (3 levels), R creates 2 dummies; for binary categoricals like `gender`, it creates 1 dummy.

```{r prepare-train}
train <- train_clean
train$satisfaction <- factor(train$satisfaction)
```

## Feature Scaling

Before fitting any classification model we scale the predictor variables. This serves two purposes:

1. **Coefficient comparability** — with all numeric predictors on similar scales, the magnitude of each logistic regression coefficient directly reflects its contribution to the log-odds of satisfaction, making it easy to rank and compare predictors regardless of their original measurement units.
2. **Numerical stability** — the IRLS solver in `glm()` converges more reliably when variables do not span vastly different ranges (e.g., `flight_distance` in the thousands versus service ratings on a 0–5 scale).

We apply two scaling strategies that respect the nature of each variable group:

- **Standard scaling (z-score)** for the four *continuous* variables (`age`, `flight_distance`, `departure_delay_in_minutes`, `arrival_delay_in_minutes`). Each variable is centred at its training-set mean and divided by its training-set standard deviation. A one-unit change in a standardised coefficient corresponds to a one-SD shift in the original variable — a natural unit for comparing variables with different spreads.
- **Min-Max scaling** for the 14 *ordinal rating* variables (discrete 0–5 scale). Each rating is rescaled to $[0,\,1]$ via $(x - \min) / (\max - \min)$. This preserves the equal spacing between adjacent integer levels while placing all ratings on a common range. A coefficient of, say, 2.0 for a min-max scaled rating means that moving from the worst possible rating (0) to the best (5) changes the log-odds of satisfaction by 2.0.

All scaling parameters — means, standard deviations, minima, and maxima — are **computed from the training set only**. The test set is not scaled here; it will be preprocessed within the model evaluation pipeline.

```{r feature-scaling}
cont_vars <- c("age", "flight_distance",
               "departure_delay_in_minutes", "arrival_delay_in_minutes")

rating_vars <- c("inflight_wifi_service", "departure_arrival_time_convenient",
                 "ease_of_online_booking", "gate_location", "food_and_drink",
                 "online_boarding", "seat_comfort", "inflight_entertainment",
                 "on_board_service", "leg_room_service", "baggage_handling",
                 "checkin_service", "inflight_service", "cleanliness")

# --- Standard Scaling for Continuous Variables (parameters from train only) ---
cont_means <- colMeans(train[, cont_vars])
cont_sds   <- apply(train[, cont_vars], 2, sd)

train_scaled <- train

train_scaled[, cont_vars] <- scale(train[, cont_vars],
                                   center = cont_means, scale = cont_sds)

# --- Min-Max Scaling for Rating Variables (parameters from train only) ---
rating_mins   <- apply(train[, rating_vars], 2, min)
rating_ranges <- apply(train[, rating_vars], 2, max) - rating_mins

for (v in rating_vars) {
  train_scaled[[v]] <- (train[[v]] - rating_mins[v]) / rating_ranges[v]
}

# Verification
cat("Continuous variables — means (should be ~0) and SDs (should be ~1):\n")
round(rbind(mean = colMeans(train_scaled[, cont_vars]),
            sd   = apply(train_scaled[, cont_vars], 2, sd)), 4)
```

```{r scaling-verify-ratings}
cat("Rating variables — observed range after min-max scaling (should be [0, 1]):\n")
apply(train_scaled[, rating_vars], 2, range) %>%
  t() %>%
  as.data.frame() %>%
  setNames(c("min", "max")) %>%
  kable(digits = 4) %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

## Logistic Regression on Full Training Set

Following the approach from the course lab (Ch. 4), we fit a logistic regression using `glm()` with `family = binomial` on all available predictors. We fit on the **entire training set** to get the most reliable coefficient estimates and p-values for interpretation.

### Model Fitting

```{r logistic-full-fit}
glm_full <- glm(satisfaction ~ ., data = train_scaled, family = binomial)
summary(glm_full)
```

### Coefficient Interpretation

```{r logistic-coef-table}
coef_table <- summary(glm_full)$coefficients %>%
  as.data.frame() %>%
  rownames_to_column("Variable") %>%
  rename(Estimate = Estimate, SE = `Std. Error`,
         z_value = `z value`, p_value = `Pr(>|z|)`) %>%
  mutate(
    Odds_Ratio = round(exp(Estimate), 4),
    Significant_0.05 = ifelse(p_value < 0.05, "Yes", "No"),
    p_value = signif(p_value, 4)
  ) %>%
  arrange(p_value)

coef_table %>%
  kable(digits = 4) %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE) %>%
  scroll_box(height = "400px")
```

In logistic regression, the coefficients are on the log-odds scale: a positive coefficient means that increasing that predictor raises the log-odds (and therefore the probability) of the positive class ("satisfied"), while a negative coefficient decreases it. The **odds ratio** (`exp(Estimate)`) gives a more intuitive interpretation: an odds ratio greater than 1 means the predictor increases the odds of satisfaction, less than 1 decreases them.

Because predictors are now scaled, coefficients are directly comparable in magnitude across all variables: **continuous variable** coefficients represent the log-odds change per one standard deviation; **rating variable** coefficients represent the change when moving from the minimum to the maximum of the full rating scale (0 → 5); **categorical variable** coefficients are unchanged, representing the difference from the reference category holding all others fixed.

Key observations from the coefficient table:

- `Customer_type` is likely the most influential predictor: being a Loyal Customer increases the odds of satisfaction by a factor of 8.54, keeping all other variables constant.
-`Type_of_travel` has a strong negative effect: passengers traveling for Personal Travel have substantially lower odds of satisfaction (OR = 0.067) compared to business travelers.
-`Class` also plays a significant role: compared to the reference category (Business class), passengers in Economy (OR = 0.45) and Economy Plus (OR = 0.37) are significantly less likely to be satisfied.

- The **Null deviance** vs **Residual deviance** comparison in the summary output shows a substantial reduction, confirming that the predictors collectively explain a meaningful portion of the variation in satisfaction.
- Some variables (e.g., `gate_location`, `gender`) have p-values well above 0.05, suggesting they contribute little after accounting for the other predictors. These are candidates for removal via feature selection.


## Feature Selection: Forward Stepwise via `step()`

The full model includes all predictors, but some contribute little after controlling for the others. Since the dataset has 23 dummy-encoded predictors (exceeding the 15-variable limit of `bestglm`) and ~100k observations, we use R's `step()` function with forward selection to identify the best model under both AIC and BIC.

**How `step()` works:** Starting from the null (intercept-only) model, forward selection evaluates every possible single-variable addition at each iteration and picks the one that most improves the information criterion. The process stops when no addition improves the criterion.

**Information criteria recap:**

- **AIC** = $-2\ell + 2k$: moderate penalty per parameter; tends to retain more variables.
- **BIC** = $-2\ell + k\log n$: stronger penalty (for our training size of ~100k, $\log n \approx 11.5$); favours more parsimonious models.

### Forward Selection with AIC

```{r step-forward-aic, cache=TRUE}
full_mod <- glm(satisfaction ~ ., data = train_scaled, family = binomial)
null_mod <- glm(satisfaction ~ 1, data = train_scaled, family = binomial)

# Forward selection using AIC (default k = 2)
step_fwd_aic <- step(null_mod,
                     scope = list(lower = null_mod, upper = full_mod),
                     direction = "forward",
                     trace = 0)

cat("=== Forward Selection — AIC ===\n")
cat("Predictors selected:", length(coef(step_fwd_aic)) - 1, "\n")
cat("AIC:", round(AIC(step_fwd_aic), 2), "\n")
cat("BIC:", round(BIC(step_fwd_aic), 2), "\n\n")
summary(step_fwd_aic)
```

### Forward Selection with BIC

```{r step-forward-bic, cache=TRUE}
# Forward selection using BIC (k = log(n))
step_fwd_bic <- step(null_mod,
                     scope = list(lower = null_mod, upper = full_mod),
                     direction = "forward",
                     k = log(nrow(train_scaled)),
                     trace = 0)

cat("=== Forward Selection — BIC ===\n")
cat("Predictors selected:", length(coef(step_fwd_bic)) - 1, "\n")
cat("AIC:", round(AIC(step_fwd_bic), 2), "\n")
cat("BIC:", round(BIC(step_fwd_bic), 2), "\n\n")
summary(step_fwd_bic)
```
The BIC-selected model retains 19 predictors, dropping `gate_location`, `food_and_drink`, `genderMale`, and `flight_distance`. Comparing with the full model, the coefficient estimates for the retained variables change only marginally, confirming that the dropped variables were neither confounders nor suppressors. The key coefficients in the reduced model remain consistent with our earlier interpretation: `customer_typeLoyal Customer` carries a large positive effect (high odds of satisfaction), `type_of_travelPersonal Travel` a large negative effect, and the service ratings `online_boarding`, `inflight_entertainment`, and `seat_comfort` retain their strong positive associations with satisfaction. The BIC model achieves comparable deviance to the AIC model with four fewer parameters, confirming it as the more parsimonious choice without sacrificing explanatory power.

### Comparison of Selected Models

```{r step-comparison}
aic_vars <- names(coef(step_fwd_aic))[-1]
bic_vars <- names(coef(step_fwd_bic))[-1]

cat("=== Forward AIC model:", length(aic_vars), "predictors ===\n")
cat(paste(aic_vars, collapse = ", "), "\n\n")

cat("=== Forward BIC model:", length(bic_vars), "predictors ===\n")
cat(paste(bic_vars, collapse = ", "), "\n\n")

# Variables in AIC but not BIC
dropped <- setdiff(aic_vars, bic_vars)
if (length(dropped) > 0) {
  cat("Dropped by BIC (extra penalty):", paste(dropped, collapse = ", "), "\n")
} else {
  cat("Both criteria selected the same variables.\n")
}
```

The AIC forward selection retained all 23 predictors — the penalty was not strong enough to exclude any variable given the large sample size. The BIC model, with its heavier penalty ($\log n \approx 11.5$ vs 2 for AIC), dropped four predictors: `gate_location`, `food_and_drink`, `genderMale`, and `flight_distance`. These are precisely the variables that showed high p-values in the full logistic regression, confirming that they contribute negligible discriminative power after accounting for the other predictors. We adopt the **BIC-selected 19-predictor model** as our reduced model going forward, as it achieves nearly the same fit with fewer parameters.




## Linear Discriminant Analysis (LDA)

LDA finds a linear combination of predictors that maximises the ratio of between-class variance to within-class variance — i.e., Fisher's criterion. For a binary problem ($K = 2$), this produces a single discriminant axis, LD1. LDA assumes that both classes share a common covariance matrix $\Sigma$, so the log-ratio of the class-conditional densities simplifies to a linear function of the predictors, yielding a linear decision boundary.

Formally, LDA classifies an observation $\mathbf{x}$ to the class $k$ that maximises the discriminant score:

$$\delta_k(\mathbf{x}) = \mathbf{x}^\top \Sigma^{-1} \boldsymbol{\mu}_k - \tfrac{1}{2} \boldsymbol{\mu}_k^\top \Sigma^{-1} \boldsymbol{\mu}_k + \log \pi_k$$

where $\boldsymbol{\mu}_k$ is the class mean vector, $\Sigma$ is the pooled within-class covariance, and $\pi_k$ is the prior probability for class $k$.

### LDA on Full Model

```{r lda-full}
lda_full <- lda(satisfaction ~ ., data = train_scaled)
lda_full
```

### LD1 Scaling (Discriminant Coefficients)

The LD1 scaling vector defines the linear combination of predictors that forms the discriminant axis. Each coefficient indicates how much each predictor contributes to the discriminant axis — because predictors have been standardised, the magnitudes are directly comparable, with larger absolute values signalling stronger contribution to class separation.

```{r lda-full-coefficients, fig.height=7}
# Extract and sort LD1 coefficients
lda_coefs <- data.frame(
  Variable = rownames(lda_full$scaling),
  LD1      = round(lda_full$scaling[, 1], 4)
) %>%
  arrange(desc(abs(LD1)))

# Table of all coefficients
lda_coefs %>%
  kable(row.names = FALSE, col.names = c("Variable", "LD1 Coefficient"),
        caption = "LD1 Scaling Coefficients (sorted by |coefficient|)") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

```{r lda-coef-barplot, fig.height=7}
# Visual: horizontal bar plot of LD1 coefficients
lda_coefs %>%
  mutate(Variable = factor(Variable, levels = rev(Variable))) %>%
  ggplot(aes(x = LD1, y = Variable, fill = LD1 > 0)) +
  geom_col(alpha = 0.8, show.legend = FALSE) +
  scale_fill_manual(values = c("TRUE" = "#3498DB", "FALSE" = "#E74C3C")) +
  geom_vline(xintercept = 0, linewidth = 0.5) +
  labs(title = "LDA Scaling Coefficients (LD1)",
       subtitle = "Blue = positive (pushes toward satisfied), Red = negative (pushes toward dissatisfied)",
       x = "LD1 Coefficient", y = NULL) +
  theme_minimal(base_size = 11) +
  theme(plot.subtitle = element_text(color = "grey40"))
```

**Interpretation of the scaling vector.**  The LD1 scaling vector defines the linear combination of predictors that forms the discriminant axis — the direction in feature space that best separates the two classes. The three strongest drivers of satisfaction are online_boarding (1.613), inflight_wifi_service (1.365), and customer_type: Loyal Customer (1.252), while type_of_travel: Personal Travel (-1.544) and ease_of_online_booking (-0.819) are the strongest drivers of dissatisfaction. Predictors such as genderMale (0.029) and flight_distance (0.005) contribute negligibly to the discriminant axis.

**Connection with BIC feature selection.** Reassuringly, the variables dropped by BIC in the logistic regression step — `gate_location`, `food_and_drink`, `genderMale`, and `flight_distance` — are precisely those that receive the smallest LD1 coefficients in absolute value. For instance, `genderMale` (0.029) and `flight_distance` (0.005) are the two weakest contributors to the discriminant axis, while `gate_location` and `food_and_drink` also rank near the bottom. This convergence across two independent methods (penalised likelihood selection and Fisher's discriminant) strengthens our confidence that these variables are genuinely uninformative for predicting satisfaction once the remaining predictors are accounted for.

**Group means.** We observe that predictors with larger differences in group means tend to receive larger LD1 coefficients in absolute value. For example, online_boarding (group means: 0.531 vs 0.805, difference: 0.274) has the largest positive coefficient (1.613), and type_of_travel: Personal Travel (0.492 vs 0.073, difference: 0.419) has the largest negative coefficient (-1.544). Conversely, predictors with nearly identical group means across classes receive near-zero coefficients, such as genderMale (0.487 vs 0.500, difference: 0.013, coefficient: 0.029) and departure_delay_in_minutes (0.044 vs -0.058, difference: 0.102, coefficient: 0.083). This is consistent with the LDA closed-form solution where between-class mean differences drive the discriminant axis.

### LD1 Score Distribution

Since we have a binary classification problem, LDA produces a single discriminant axis (LD1). The LD1 scores represent each observation's position along the direction that maximally separates the two classes. Visualising these scores as a histogram with kernel density estimation (KDE) overlays provides a direct view of how well LDA discriminates between satisfied and dissatisfied passengers.

```{r lda-full-ld1-histogram, fig.height=6}
lda_full_pred <- predict(lda_full)

ld1_df <- data.frame(
  LD1   = lda_full_pred$x[, 1],
  Class = train_scaled$satisfaction
)

# Compute the LDA decision threshold (midpoint of group means on LD1)
#group_means_ld1 <- tapply(ld1_df$LD1, ld1_df$Class, mean)
#lda_threshold <- mean(group_means_ld1)

ggplot(ld1_df, aes(x = LD1, fill = Class)) +
  geom_histogram(aes(y = after_stat(density)), bins = 80,
                 alpha = 0.4, position = "identity", color = "white", linewidth = 0.1) +
  geom_density(aes(color = Class), linewidth = 1, fill = NA) +
#  geom_vline(xintercept = lda_threshold, linetype = "dashed", color = "grey30", linewidth = 0.7) +
#  annotate("text", x = lda_threshold, y = Inf, label = "Decision\nThreshold",
#           vjust = 1.5, hjust = -0.1, size = 3, color = "grey30") +
  scale_fill_manual(values = c("neutral or dissatisfied" = "#E74C3C",
                                "satisfied" = "#3498DB")) +
  scale_color_manual(values = c("neutral or dissatisfied" = "#C0392B",
                                 "satisfied" = "#2980B9")) +
  labs(title = "LDA: Distribution of LD1 Scores by Class",
       subtitle = "Histogram with KDE overlay ",
       x = "LD1 Score", y = "Density",
       fill = "Class", color = "Class") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom",
        plot.subtitle = element_text(color = "grey40"))
```

The two classes are well separated along LD1, with satisfied passengers concentrating around positive scores and dissatisfied passengers around negative scores. The overlap region near zero represents observations where the two classes are hardest to distinguish, and corresponds to the highest classification uncertainty under the model. The skewness visible in the satisfied class suggests the normality assumption may be imperfect, which motivates also fitting QDA as a comparison.


## Naive Bayes on Full Training Data

Before proceeding to dimensionality reduction, we fit a Naive Bayes classifier with default hyperparameters on the full training set. The purpose is not model selection but exploratory interpretation: Naive Bayes estimates class-conditional distributions for each predictor independently, and the resulting summary gives us a concise snapshot of how each feature's mean and spread differ between satisfied and dissatisfied passengers.

```{r naive-bayes-full-exploration}
library(e1071)

nb_explore <- naiveBayes(satisfaction ~ ., data = train_scaled)
nb_explore
```

**Interpretation of the Naive Bayes summary.**  The output shows the class-conditional means and standard deviations for each numeric predictor, as well as the conditional class probabilities for each level of the categorical predictors. Key observations:

- The **a-priori probabilities** reflect the class balance we observed earlier (~56.7% dissatisfied, ~43.3% satisfied).
- Among the continuous variables, `online_boarding` and `inflight_entertainment` show the largest separation in class-conditional means — satisfied passengers rate these much higher on average — which aligns with the LDA and logistic regression findings.
- Variables like `departure_delay_in_minutes` and `arrival_delay_in_minutes` have very similar means across both classes, confirming that delays contribute little to the satisfaction prediction.
- For categorical predictors, `type_of_travel` and `customer_type` show large differences in conditional probabilities between classes, again consistent with the strong effects identified by logistic regression and LDA.

The Naive Bayes assumption of conditional independence between predictors is clearly violated (e.g., departure and arrival delay are almost perfectly correlated), so we do not expect Naive Bayes to be the top-performing classifier. However, its simplicity and interpretability make it a useful baseline for comparison.


## Principal Component Analysis (PCA)

Before fitting PCA-based classifiers, we apply PCA to the numeric predictor matrix to reduce dimensionality. PCA finds orthogonal directions of maximum variance in the data, which can help mitigate multicollinearity and reduce the number of parameters that need to be estimated — particularly relevant for QDA, which estimates a full covariance matrix per class.

We apply PCA to the **model matrix** (with dummy-encoded categoricals) to ensure all predictors are numeric.

```{r pca-fit}
# Build numeric model matrix (same as used for step selection)
train_mm_pca <- as.data.frame(
  model.matrix(satisfaction ~ ., data = train_scaled)[, -1]
)

# PCA on the predictor matrix (already scaled)
pca_fit <- prcomp(train_mm_pca, center = TRUE, scale. = TRUE)

# Variance explained
pca_var <- summary(pca_fit)
pca_var
```

### Scree Plot and Cumulative Variance

```{r pca-scree, fig.height=5}
var_explained <- pca_var$importance[2, ]  # Proportion of variance
cum_var       <- pca_var$importance[3, ]  # Cumulative proportion

pca_df <- data.frame(
  PC = seq_along(var_explained),
  Variance = var_explained,
  Cumulative = cum_var
)

# Number of PCs for 90% variance
n_pc_90 <- min(which(cum_var >= 0.90))
cat("PCs needed for 90% variance:", n_pc_90, "\n")

p1 <- ggplot(pca_df, aes(x = PC, y = Variance)) +
  geom_col(fill = "steelblue", alpha = 0.7) +
  geom_line(color = "darkblue", linewidth = 0.8) +
  geom_point(color = "darkblue", size = 2) +
  labs(title = "Scree Plot", x = "Principal Component", y = "Proportion of Variance") +
  scale_x_continuous(breaks = 1:length(var_explained)) +
  theme_minimal()

p2 <- ggplot(pca_df, aes(x = PC, y = Cumulative)) +
  geom_line(color = "tomato", linewidth = 0.8) +
  geom_point(color = "tomato", size = 2) +
  geom_hline(yintercept = 0.90, linetype = "dashed", color = "grey40") +
  geom_vline(xintercept = n_pc_90, linetype = "dashed", color = "grey40") +
  annotate("text", x = n_pc_90 + 0.5, y = 0.85,
           label = paste0(n_pc_90, " PCs = 90%"), hjust = 0, size = 3.5) +
  labs(title = "Cumulative Variance Explained",
       x = "Principal Component", y = "Cumulative Proportion") +
  scale_x_continuous(breaks = 1:length(var_explained)) +
  theme_minimal()

grid.arrange(p1, p2, ncol = 2)
```

The scree plot shows how much variance each principal component captures. The cumulative plot identifies how many components are needed to reach the 90% threshold. We retain the first `r n_pc_90` principal components for the PCA-based classifiers.

Note: We do not produce PCA-based decision boundary plots for LDA and QDA. Projecting the data onto just two principal components and then fitting a 2D classifier discards the information from all other components, which makes the resulting boundaries misleading. The LD1 histogram above already provides a more faithful representation of how LDA separates the classes using the full predictor space. For QDA, which does not produce a single discriminant axis, we focus instead on comparing group means and on predictive performance metrics in the evaluation section below.

# Model Evaluation and Selection

We now proceed to a systematic comparison of five classification methods — Logistic Regression, LDA, QDA, KNN, and Naive Bayes — each fitted with **default hyperparameters** (no tuning) across three predictor (data version) configurations: the **Full** model (all predictors), the **BIC-reduced** model (19 predictors from forward stepwise selection), and the **PCA-transformed** data (principal components explaining 90% of variance). This yields a complete 5 × 3 = 15 model–data version grid.

**Important methodological note on data leakage.** To avoid data leakage, we perform all preprocessing — imputation and scaling — within each evaluation step. All imputation models and scaling parameters are fitted exclusively on the training portion and applied to the validation/test portions. This ensures that no information from the evaluation data leaks into the preprocessing.

## Train–Validation–Test Split

We use the full `train.csv` (80% of the original data) as the training set. The `test.csv` (20% of the original data) is split 50/50 into a **validation set** (10% of total) for model selection and a **held-out test set** (10% of total) for final evaluation. This yields an 80–10–10 split overall. All imputation and scaling parameters are estimated from the training set only and applied to all sets.

```{r train-val-test-split}
set.seed(42)
test_idx <- sample(nrow(test_with_na), size = round(0.5 * nrow(test_with_na)))
val_raw  <- test_with_na[-test_idx, ]
test_holdout <- test_with_na[test_idx, ]

fit_raw <- train_with_na

cat("Training set:", nrow(fit_raw), "observations\n")
cat("Validation set:", nrow(val_raw), "observations\n")
cat("Held-out test set:", nrow(test_holdout), "observations\n")
cat("Class balance (train):", round(prop.table(table(fit_raw$satisfaction)) * 100, 1), "%\n")
cat("Class balance (val):", round(prop.table(table(val_raw$satisfaction)) * 100, 1), "%\n")
cat("Class balance (test):", round(prop.table(table(test_holdout$satisfaction)) * 100, 1), "%\n")
```

### Imputation and Scaling on the Training Set (No Leakage)

We fit the imputation model and compute all scaling parameters from the fitting set only, then apply them to both sets.

```{r leakage-free-preprocessing}
# --- 1. Imputation: fit on fitting set, apply to both ---
fit_reg <- fit_raw %>%
  dplyr::select(departure_delay_in_minutes, arrival_delay_in_minutes) %>%
  na.omit()

delay_model_fit <- lm(arrival_delay_in_minutes ~ departure_delay_in_minutes,
                      data = fit_reg)

train_fit <- fit_raw
train_val <- val_raw

train_fit$arrival_delay_in_minutes[is.na(train_fit$arrival_delay_in_minutes)] <-
  predict(delay_model_fit,
          newdata = train_fit[is.na(train_fit$arrival_delay_in_minutes), ])

train_val$arrival_delay_in_minutes[is.na(train_val$arrival_delay_in_minutes)] <-
  predict(delay_model_fit,
          newdata = train_val[is.na(train_val$arrival_delay_in_minutes), ])

cat("NAs remaining — fit:", sum(is.na(train_fit$arrival_delay_in_minutes)),
    " val:", sum(is.na(train_val$arrival_delay_in_minutes)), "\n")

# --- 2. Convert target to factor ---
train_fit$satisfaction <- factor(train_fit$satisfaction)
train_val$satisfaction <- factor(train_val$satisfaction)

# --- 3. Scaling: compute parameters from fitting set only ---
cont_vars <- c("age", "flight_distance",
               "departure_delay_in_minutes", "arrival_delay_in_minutes")

rating_vars <- c("inflight_wifi_service", "departure_arrival_time_convenient",
                 "ease_of_online_booking", "gate_location", "food_and_drink",
                 "online_boarding", "seat_comfort", "inflight_entertainment",
                 "on_board_service", "leg_room_service", "baggage_handling",
                 "checkin_service", "inflight_service", "cleanliness")

# Standard scaling for continuous vars
fit_cont_means <- colMeans(train_fit[, cont_vars])
fit_cont_sds   <- apply(train_fit[, cont_vars], 2, sd)

train_fit[, cont_vars] <- scale(train_fit[, cont_vars],
                                 center = fit_cont_means, scale = fit_cont_sds)
train_val[, cont_vars] <- scale(train_val[, cont_vars],
                                 center = fit_cont_means, scale = fit_cont_sds)

# Min-max scaling for rating vars
fit_rating_mins   <- apply(train_fit[, rating_vars], 2, min)
fit_rating_ranges <- apply(train_fit[, rating_vars], 2, max) - fit_rating_mins

for (v in rating_vars) {
  train_fit[[v]] <- (train_fit[[v]] - fit_rating_mins[v]) / fit_rating_ranges[v]
  train_val[[v]] <- (train_val[[v]] - fit_rating_mins[v]) / fit_rating_ranges[v]
}

cat("Preprocessing complete .\n")
```

### Preparing the Three Predictor Sets

We construct three versions of the predictor matrix. For the PCA version, the principal component transformation is fitted on the **training (fitting) set only** and then applied to the validation set using `predict()`, so that the validation set contributes no information to the rotation — avoiding data leakage in the same way we handled imputation and scaling above.

```{r prepare-predictor-sets}
# --- BIC-reduced formula ---
bic_formula <- formula(step_fwd_bic)

# --- PCA predictor sets ---
# Fit PCA on the fitting set only; apply the learned rotation to the validation set
train_fit_mm <- as.data.frame(model.matrix(satisfaction ~ ., data = train_fit)[, -1])
train_val_mm <- as.data.frame(model.matrix(satisfaction ~ ., data = train_val)[, -1])

pca_fit_only <- prcomp(train_fit_mm, center = TRUE, scale. = TRUE)
cum_var_fit  <- cumsum(pca_fit_only$sdev^2 / sum(pca_fit_only$sdev^2))
n_pc_90_fit  <- min(which(cum_var_fit >= 0.90))

# Transform both sets using fitting-set PCA
train_fit_pca <- as.data.frame(predict(pca_fit_only, train_fit_mm)[, 1:n_pc_90_fit])
train_val_pca <- as.data.frame(predict(pca_fit_only, train_val_mm)[, 1:n_pc_90_fit])
train_fit_pca$satisfaction <- train_fit$satisfaction
train_val_pca$satisfaction <- train_val$satisfaction

cat("PCA components retained (90% variance on fitting set):", n_pc_90_fit, "\n")
```
```{r}
bic_formula
```
## Initial Model Comparison (Default Hyperparameters)

All models are fitted with **default hyperparameters** (no tuning). We report confusion matrices and classification metrics on the held-out validation set.

```{r fit-all-models-default, cache=TRUE}
library(class)   # KNN
library(e1071)   # Naive Bayes
# =============================================
# Fit all models and get predicted probabilities for "satisfied"
# =============================================

val_labels <- train_val$satisfaction

# --- 1. Logistic Regression (Full) ---
glm_full_fit <- glm(satisfaction ~ ., data = train_fit, family = binomial)
prob_glm_full <- predict(glm_full_fit, train_val, type = "response")

# --- 2. Logistic Regression (BIC reduced) ---
glm_bic_fit <- glm(bic_formula, data = train_fit, family = binomial)
prob_glm_bic <- predict(glm_bic_fit, train_val, type = "response")

# --- 3. Logistic Regression (PCA) ---
glm_pca_fit <- glm(satisfaction ~ ., data = train_fit_pca, family = binomial)
prob_glm_pca <- predict(glm_pca_fit, train_val_pca, type = "response")

# --- 4. LDA (Full) ---
lda_full_fit <- lda(satisfaction ~ ., data = train_fit)
prob_lda_full <- predict(lda_full_fit, train_val)$posterior[, "satisfied"]

# --- 5. LDA (BIC reduced) ---
lda_bic_fit <- lda(bic_formula, data = train_fit)
prob_lda_bic <- predict(lda_bic_fit, train_val)$posterior[, "satisfied"]

# --- 6. LDA (PCA) ---
lda_pca_fit <- lda(satisfaction ~ ., data = train_fit_pca)
prob_lda_pca <- predict(lda_pca_fit, train_val_pca)$posterior[, "satisfied"]

# --- 7. QDA (Full) ---
qda_full_fit <- qda(satisfaction ~ ., data = train_fit)
prob_qda_full <- predict(qda_full_fit, train_val)$posterior[, "satisfied"]

# --- 8. QDA (BIC reduced) ---
qda_bic_fit <- qda(bic_formula, data = train_fit)
prob_qda_bic <- predict(qda_bic_fit, train_val)$posterior[, "satisfied"]

# --- 9. QDA (PCA) ---
qda_pca_fit <- qda(satisfaction ~ ., data = train_fit_pca)
prob_qda_pca <- predict(qda_pca_fit, train_val_pca)$posterior[, "satisfied"]

# --- 10. KNN (Full, BIC, PCA) — default k=5 ---
train_fit_knn <- as.data.frame(model.matrix(satisfaction ~ ., data = train_fit)[, -1])
train_val_knn <- as.data.frame(model.matrix(satisfaction ~ ., data = train_val)[, -1])
train_fit_knn_bic <- as.data.frame(model.matrix(bic_formula, data = train_fit)[, -1])
train_val_knn_bic <- as.data.frame(model.matrix(bic_formula, data = train_val)[, -1])

knn_pred_full <- knn(train_fit_knn, train_val_knn,
                     cl = train_fit$satisfaction, k = 5, prob = TRUE)
knn_prob <- attr(knn_pred_full, "prob")
prob_knn_full <- ifelse(knn_pred_full == "satisfied", knn_prob, 1 - knn_prob)

knn_pred_bic <- knn(train_fit_knn_bic, train_val_knn_bic,
                    cl = train_fit$satisfaction, k = 5, prob = TRUE)
knn_prob_b <- attr(knn_pred_bic, "prob")
prob_knn_bic <- ifelse(knn_pred_bic == "satisfied", knn_prob_b, 1 - knn_prob_b)

knn_pred_pca <- knn(train_fit_pca[, -ncol(train_fit_pca)],
                    train_val_pca[, -ncol(train_val_pca)],
                    cl = train_fit$satisfaction, k = 5, prob = TRUE)
knn_prob_p <- attr(knn_pred_pca, "prob")
prob_knn_pca <- ifelse(knn_pred_pca == "satisfied", knn_prob_p, 1 - knn_prob_p)

# --- 11. Naive Bayes (Full, BIC, PCA) ---
nb_full_fit <- naiveBayes(satisfaction ~ ., data = train_fit)
prob_nb_full <- predict(nb_full_fit, train_val, type = "raw")[, "satisfied"]

nb_bic_fit <- naiveBayes(bic_formula, data = train_fit)
prob_nb_bic <- predict(nb_bic_fit, train_val, type = "raw")[, "satisfied"]

nb_pca_fit <- naiveBayes(satisfaction ~ ., data = train_fit_pca)
prob_nb_pca <- predict(nb_pca_fit, train_val_pca, type = "raw")[, "satisfied"]

cat("All models fitted successfully.\n")
```

### Validation Set Metrics

```{r initial-metrics-table}
# Collect all probabilities
all_probs <- list(
  "GLM Full"  = prob_glm_full,  "GLM BIC"  = prob_glm_bic,  "GLM PCA"  = prob_glm_pca,
  "LDA Full"  = prob_lda_full,  "LDA BIC"  = prob_lda_bic,  "LDA PCA"  = prob_lda_pca,
  "QDA Full"  = prob_qda_full,  "QDA BIC"  = prob_qda_bic,  "QDA PCA"  = prob_qda_pca,
  "KNN5 Full" = prob_knn_full,  "KNN5 BIC" = prob_knn_bic,  "KNN5 PCA" = prob_knn_pca,
  "NB Full"   = prob_nb_full,   "NB BIC"   = prob_nb_bic,   "NB PCA"   = prob_nb_pca
)

# Helper: compute metrics
compute_metrics <- function(probs, labels, threshold = 0.5) {
  pred_class <- factor(ifelse(probs >= threshold, "satisfied", "neutral or dissatisfied"),
                       levels = levels(labels))
  cm <- confusionMatrix(pred_class, labels, positive = "satisfied")
  roc_obj <- roc(labels, probs,
                 levels = c("neutral or dissatisfied", "satisfied"),
                 direction = "<", quiet = TRUE)
  prec <- cm$byClass["Precision"]
  rec  <- cm$byClass["Recall"]
  f1   <- ifelse(prec + rec > 0, 2 * prec * rec / (prec + rec), 0)
  data.frame(
    Accuracy  = round(cm$overall["Accuracy"], 4),
    Precision = round(prec, 4),
    Recall    = round(rec, 4),
    F1        = round(f1, 4),
    AUC       = round(as.numeric(auc(roc_obj)), 4)
  )
}

# Build metrics table
metrics_list <- lapply(names(all_probs), function(nm) {
  m <- compute_metrics(all_probs[[nm]], val_labels)
  m$Model <- nm
  m
})
metrics_df <- bind_rows(metrics_list) %>%
  dplyr::select(Model, Accuracy, Precision, Recall, F1, AUC) %>%
  arrange(desc(AUC))

metrics_df %>%
  kable(row.names = FALSE,
        caption = "Initial Model Comparison — Validation Set Metrics (Default Hyperparameters)") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

### Selecting the Best Model–Data Version Combinations

We use **AUC (Area Under the ROC Curve)** as the primary metric for model selection. This choice is motivated by two considerations. First, the dataset exhibits a moderate class imbalance (~57% dissatisfied vs ~43% satisfied), which means that raw accuracy can be misleading — a model that simply predicts the majority class would achieve ~57% accuracy without learning anything useful. AUC, by contrast, evaluates the model's ability to discriminate between the two classes across *all possible classification thresholds*, making it robust to class imbalance. Second, AUC directly measures the probability that a randomly chosen satisfied passenger receives a higher predicted probability than a randomly chosen dissatisfied passenger — a threshold-free summary of overall separability that is particularly informative when the optimal decision threshold is not known in advance.

Rather than simply picking the top 4 individual configurations, we identify the **best 2 models** and the **best 2 data versions** and form their 4 combinations. This structured approach lets us disentangle the effect of the model choice from the effect of the predictor set, and ensures we carry forward configurations that are strong on both dimensions.

```{r select-best-combinations}
# Parse model name into method + data version
metrics_df <- metrics_df %>%
  mutate(
    Method      = sub(" .*", "", Model),
    DataVersion = sub(".* ", "", Model)
  )

# --- Best 2 models: average AUC across all 3 data versions ---
model_avg <- metrics_df %>%
  group_by(Method) %>%
  summarise(Mean_AUC = round(mean(AUC), 4), .groups = "drop") %>%
  arrange(desc(Mean_AUC))

cat("=== Average AUC by Model (across all data versions) ===\n")
model_avg %>%
  kable(row.names = FALSE, col.names = c("Model", "Mean AUC")) %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE)

best_2_models <- model_avg$Method[1:2]
cat("\nBest 2 models:", paste(best_2_models, collapse = ", "), "\n")

# --- Best 2 data versions: average AUC across all models ---
dv_avg <- metrics_df %>%
  group_by(DataVersion) %>%
  summarise(Mean_AUC = round(mean(AUC), 4), .groups = "drop") %>%
  arrange(desc(Mean_AUC))

cat("\n=== Average AUC by Data Version (across all models) ===\n")
dv_avg %>%
  kable(row.names = FALSE, col.names = c("Data Version", "Mean AUC")) %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE)

best_2_dv <- dv_avg$DataVersion[1:2]
cat("\nBest 2 data versions:", paste(best_2_dv, collapse = ", "), "\n")

# --- Form the 4 combinations ---
selected_configs <- expand.grid(Method = best_2_models, DataVersion = best_2_dv,
                                 stringsAsFactors = FALSE) %>%
  mutate(Model = paste(Method, DataVersion))

cat("\n=== Selected 4 Configurations (2 models x 2 data versions) ===\n")
cat(paste(selected_configs$Model, collapse = "\n"), "\n")

selected_names <- selected_configs$Model
```

### ROC Curves for Selected Configurations

```{r roc-selected, fig.height=6, fig.width=8}
roc_selected <- lapply(selected_names, function(nm) {
  roc(val_labels, all_probs[[nm]],
      levels = c("neutral or dissatisfied", "satisfied"),
      direction = "<", quiet = TRUE)
})
names(roc_selected) <- selected_names

colors_sel <- c("#E74C3C", "#3498DB", "#2ECC71", "#9B59B6")

plot(roc_selected[[1]], col = colors_sel[1], lwd = 2.5,
     main = "ROC Curves — Selected Configurations", legacy.axes = TRUE)
for (i in 2:length(roc_selected)) {
  plot(roc_selected[[i]], col = colors_sel[i], lwd = 2.5,
       add = TRUE, legacy.axes = TRUE)
}
abline(a = 0, b = 1, lty = 2, col = "grey60")
legend("bottomright",
       paste0(selected_names, " (AUC = ", round(sapply(roc_selected, auc), 4), ")"),
       col = colors_sel, lwd = 2.5, cex = 0.85)
```

### Confusion Matrices for Selected Configurations

```{r cm-selected, fig.height=6, fig.width=12}
cm_plots <- list()
for (i in seq_along(selected_names)) {
  nm <- selected_names[i]
  pred_class <- factor(ifelse(all_probs[[nm]] >= 0.5,
                              "satisfied", "neutral or dissatisfied"),
                       levels = levels(val_labels))
  cm <- table(Predicted = pred_class, Actual = val_labels)
  cm_df <- as.data.frame(cm)

  cm_plots[[i]] <- ggplot(cm_df, aes(x = Actual, y = Predicted, fill = Freq)) +
    geom_tile(color = "white", linewidth = 1.2) +
    geom_text(aes(label = format(Freq, big.mark = ",")), size = 4.5, fontface = "bold") +
    scale_fill_gradient(low = "white", high = "steelblue") +
    labs(title = nm) +
    theme_minimal(base_size = 11) +
    theme(legend.position = "none",
          axis.text.x = element_text(angle = 20, hjust = 1))
}

grid.arrange(grobs = cm_plots, ncol = 2)
```

### Metrics Summary for Selected Configurations

```{r metrics-selected}
metrics_df %>%
  filter(Model %in% selected_names) %>%
  dplyr::select(Model, Accuracy, Precision, Recall, F1, AUC) %>%
  arrange(desc(AUC)) %>%
  kable(row.names = FALSE,
        caption = "Selected Configurations — Validation Set Metrics") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

## Cross-Validation on 90% Data (Selected Configurations)

The single train–validation split above gave us an initial ranking, but a single split can be sensitive to which observations end up in the validation set — we may have simply "got lucky" with a favourable partition. To obtain a more reliable estimate of test error and to verify that our models generalise consistently, we perform **10-fold cross-validation** on the combined training + validation data (90% of the total) for the 4 selected model–data version configurations. The held-out 10% test set (`test_holdout`) remains completely untouched until final evaluation.

For KNN, which has a tuneable hyperparameter (the number of neighbours $k$), we additionally search over $k \in \{5, 10, 15, 20\}$ within the cross-validation loop.

**Pipeline approach.** A critical requirement is that *all preprocessing* — imputation, scaling, and (where applicable) PCA — must be fitted on the training folds and applied to the held-out validation fold within each iteration. If we were to impute or scale using the full dataset before splitting into folds, information from the validation fold would leak into the training folds, producing optimistically biased error estimates. We use the `recipes` package from `tidymodels` to define reusable preprocessing pipelines that enforce this discipline: each recipe is `prep()`-ed on the training folds and `bake()`-d onto the validation fold, guaranteeing that no information flows in the wrong direction.

```{r cv-setup}
# install.packages("recipes")  # uncomment on first run
library(recipes)

# Combine train (80%) + validation (10%) = 90% for cross-validation
cv_data <- bind_rows(train_with_na, val_raw)
cat("CV data size (90% of total):", nrow(cv_data), "observations\n")
cat("Held-out test size (10% of total):", nrow(test_holdout), "observations\n")

set.seed(42)
n_folds <- 10
fold_ids <- createFolds(cv_data$satisfaction, k = n_folds, returnTrain = FALSE)

# Variable groups (same as before)
cont_vars <- c("age", "flight_distance",
               "departure_delay_in_minutes", "arrival_delay_in_minutes")

rating_vars <- c("inflight_wifi_service", "departure_arrival_time_convenient",
                 "ease_of_online_booking", "gate_location", "food_and_drink",
                 "online_boarding", "seat_comfort", "inflight_entertainment",
                 "on_board_service", "leg_room_service", "baggage_handling",
                 "checkin_service", "inflight_service", "cleanliness")

cat_vars <- c("gender", "customer_type", "type_of_travel", "class")

# BIC-selected variable names (from the forward BIC step earlier)
bic_predictors <- all.vars(bic_formula)[-1]  # drop the response
cat("BIC predictors:", paste(bic_predictors, collapse = ", "), "\n")
```

### Building the Preprocessing Recipes

We define three recipes — one per data version — that encode the full preprocessing pipeline:

1. **Imputation**: predict missing `arrival_delay_in_minutes` from `departure_delay_in_minutes` using `step_impute_linear()`.
2. **Scaling**: z-score standardisation for continuous variables (`step_normalize()`), min-max scaling for ordinal ratings (`step_range()`).
3. **Dummy encoding**: create indicator variables for the categorical predictors (`step_dummy()`).
4. **PCA** (PCA version only): `step_pca()` retaining components for 90% of variance.
5. **Variable selection** (BIC version only): keep only the BIC-selected predictors (`step_select()`).

```{r define-recipes}
# --- Base recipe (shared preprocessing steps) ---
base_recipe <- function(data) {
  recipe(satisfaction ~ ., data = data) %>%
    step_impute_linear(arrival_delay_in_minutes,
                       impute_with = imp_vars(departure_delay_in_minutes)) %>%
    step_normalize(all_of(cont_vars)) %>%
    step_range(all_of(rating_vars), min = 0, max = 1) %>%
    step_dummy(all_of(cat_vars))
}

# --- Full recipe ---
make_recipe_full <- function(data) {
  base_recipe(data)
}

# --- BIC recipe: select only BIC predictors before dummy encoding ---
make_recipe_bic <- function(data) {
  recipe(satisfaction ~ ., data = data) %>%
    step_impute_linear(arrival_delay_in_minutes,
                       impute_with = imp_vars(departure_delay_in_minutes)) %>%
    step_rm(-satisfaction, -all_of(bic_predictors)) %>%
    step_normalize(all_of(intersect(cont_vars, bic_predictors))) %>%
    step_range(all_of(intersect(rating_vars, bic_predictors)), min = 0, max = 1) %>%
    step_dummy(all_of(intersect(cat_vars, bic_predictors)))
}

# --- PCA recipe: full preprocessing then PCA ---
make_recipe_pca <- function(data) {
  base_recipe(data) %>%
    step_pca(all_predictors(), threshold = 0.90)
}
```

### Cross-Validation Loop

```{r}
selected_configs
```


```{r cv-loop, cache=TRUE}
# KNN k values to try
knn_k_values <- c(5,10,15)

# Helper: get AUC
get_auc <- function(probs, labels) {
  as.numeric(auc(roc(labels, probs,
                     levels = c("neutral or dissatisfied", "satisfied"),
                     direction = "<", quiet = TRUE)))
}

# Helper: get F1 score from probabilities
get_f1 <- function(probs, labels, threshold = 0.5) {
  pred_class <- factor(ifelse(probs >= threshold, "satisfied", "neutral or dissatisfied"),
                       levels = levels(labels))
  cm <- confusionMatrix(pred_class, labels, positive = "satisfied")
  prec <- cm$byClass["Precision"]
  rec  <- cm$byClass["Recall"]
  ifelse(prec + rec > 0, 2 * prec * rec / (prec + rec), 0)
}

# Storage: AUC and F1 per fold
cv_results <- list()
cv_f1_results <- list()

for (cfg_i in seq_len(nrow(selected_configs))) {
  method <- selected_configs$Method[cfg_i]
  dv     <- selected_configs$DataVersion[cfg_i]
  cfg_name <- selected_configs$Model[cfg_i]

  cat("\n=== Configuration:", cfg_name, "===\n")

  # Choose recipe constructor
  make_recipe <- switch(dv,
    "Full" = make_recipe_full,
    "BIC"  = make_recipe_bic,
    "PCA"  = make_recipe_pca
  )

  if (method == "KNN5") {
    # KNN: store results for each k value
    for (k_val in knn_k_values) {
      result_name <- paste0(cfg_name, " (k=", k_val, ")")
      cv_results[[result_name]] <- numeric(n_folds)
      cv_f1_results[[result_name]] <- numeric(n_folds)
    }
  } else {
    cv_results[[cfg_name]] <- numeric(n_folds)
    cv_f1_results[[cfg_name]] <- numeric(n_folds)
  }

  for (f in seq_len(n_folds)) {
    idx_val <- fold_ids[[f]]
    tr_raw  <- cv_data[-idx_val, ]
    va_raw  <- cv_data[idx_val, ]

    # Prep recipe on training fold, bake both
    rec <- make_recipe(tr_raw) %>% prep(training = tr_raw)
    tr_baked <- bake(rec, new_data = NULL)  # NULL = return training data
    va_baked <- bake(rec, new_data = va_raw)

    va_lab <- va_baked$satisfaction

    tryCatch({
      if (method == "GLM") {
        fit <- glm(satisfaction ~ ., data = tr_baked, family = binomial)
        probs <- predict(fit, va_baked, type = "response")
        cv_results[[cfg_name]][f] <- get_auc(probs, va_lab)
        cv_f1_results[[cfg_name]][f] <- get_f1(probs, va_lab)

      } else if (method == "LDA") {
        fit <- lda(satisfaction ~ ., data = tr_baked)
        probs <- predict(fit, va_baked)$posterior[, "satisfied"]
        cv_results[[cfg_name]][f] <- get_auc(probs, va_lab)
        cv_f1_results[[cfg_name]][f] <- get_f1(probs, va_lab)

      } else if (method == "QDA") {
        fit <- qda(satisfaction ~ ., data = tr_baked)
        probs <- predict(fit, va_baked)$posterior[, "satisfied"]
        cv_results[[cfg_name]][f] <- get_auc(probs, va_lab)
        cv_f1_results[[cfg_name]][f] <- get_f1(probs, va_lab)

      } else if (method == "NB") {
        fit <- naiveBayes(satisfaction ~ ., data = tr_baked)
        probs <- predict(fit, va_baked, type = "raw")[, "satisfied"]
        cv_results[[cfg_name]][f] <- get_auc(probs, va_lab)
        cv_f1_results[[cfg_name]][f] <- get_f1(probs, va_lab)

      } else if (method == "KNN5") {
        # Prepare numeric matrices for knn()
        tr_x <- tr_baked %>% dplyr::select(-satisfaction) %>% as.data.frame()
        va_x <- va_baked %>% dplyr::select(-satisfaction) %>% as.data.frame()
        tr_y <- tr_baked$satisfaction

        for (k_val in knn_k_values) {
          result_name <- paste0(cfg_name, " (k=", k_val, ")")
          knn_pred <- knn(tr_x, va_x, cl = tr_y, k = k_val, prob = TRUE)
          knn_pr   <- attr(knn_pred, "prob")
          probs    <- ifelse(knn_pred == "satisfied", knn_pr, 1 - knn_pr)
          cv_results[[result_name]][f] <- get_auc(probs, va_lab)
          cv_f1_results[[result_name]][f] <- get_f1(probs, va_lab)
        }
      }
    }, error = function(e) {
      cat("  Error in fold", f, ":", conditionMessage(e), "\n")
    })
  }
  cat("  Done.\n")
}

cat("\n10-fold cross-validation complete.\n")
```

### Cross-Validation AUC Results

```{r cv-results-table}
cv_summary <- data.frame(
  Configuration = names(cv_results),
  Mean_AUC      = round(sapply(cv_results, mean, na.rm = TRUE), 4),
  SD_AUC        = round(sapply(cv_results, sd, na.rm = TRUE), 4),
  Mean_F1       = round(sapply(cv_f1_results, mean, na.rm = TRUE), 4),
  SD_F1         = round(sapply(cv_f1_results, sd, na.rm = TRUE), 4)
) %>%
  arrange(desc(Mean_AUC))

cv_summary %>%
  kable(row.names = FALSE,
        col.names = c("Configuration", "Mean AUC", "SD AUC", "Mean F1", "SD F1"),
        caption = "10-Fold Cross-Validation Results (90% Data: Train + Validation)") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

```{r cv-boxplot, fig.height=6, fig.width=10}
cv_long <- stack(cv_results) %>%
  rename(AUC = values, Configuration = ind)

cv_long$Configuration <- factor(cv_long$Configuration,
                                 levels = rev(cv_summary$Configuration))

ggplot(cv_long, aes(x = Configuration, y = AUC, fill = Configuration)) +
  geom_boxplot(alpha = 0.7, show.legend = FALSE) +
  coord_flip() +
  labs(title = "10-Fold Cross-Validation: AUC Distribution",
       subtitle = "KNN configurations shown with different k values",
       x = "", y = "AUC") +
  theme_minimal(base_size = 11)
```

### Best Configuration

```{r best-cv-config}
best_cv <- cv_summary$Configuration[1]
best_cv_auc <- cv_summary$Mean_AUC[1]
best_cv_sd  <- cv_summary$SD_AUC[1]

cat("=== Best Configuration by Mean 10-Fold CV AUC ===\n")
cat("Configuration:", best_cv, "\n")
cat("Mean AUC:", best_cv_auc, "(±", best_cv_sd, ")\n\n")

# If best is a KNN variant, extract the best k
if (grepl("KNN", best_cv)) {
  best_k <- as.numeric(gsub(".*k=(\\d+).*", "\\1", best_cv))
  cat("Best k for KNN:", best_k, "\n")
}

cat("\nFull ranking:\n")
for (i in seq_len(nrow(cv_summary))) {
  cat(sprintf("  %d. %s — Mean AUC: %.4f (± %.4f)\n",
              i, cv_summary$Configuration[i],
              cv_summary$Mean_AUC[i], cv_summary$SD_AUC[i]))
}
```
## Logistic Regression with L1 and L2 penalty
```{r}
#install.packages("tidymodels")
library(tidymodels)
library(glmnet)
library(pROC)
library(dplyr)
library(ggplot2)
library(tibble)

#--------------------------------------------------
# AUC helper
#--------------------------------------------------
get_auc <- function(probs, labels) {
  as.numeric(auc(
    roc(labels, probs,
        levels = c("neutral or dissatisfied", "satisfied"),
        direction = "<",
        quiet = TRUE)
  ))
}

# glmnet needs 0/1 outcome for binomial
to_binary01 <- function(y) {
  ifelse(y == "satisfied", 1, 0)
}

#--------------------------------------------------
# CV over a lambda grid for penalized logistic regression
# alpha = 0 -> ridge
# alpha = 1 -> lasso
#--------------------------------------------------
cv_glmnet_grid <- function(data,
                           fold_ids,
                           make_recipe,
                           lambda_grid,
                           alpha_value,
                           model_name = "ridge") {

  n_folds <- length(fold_ids)

  # rows = lambdas, cols = folds
  auc_mat <- matrix(NA_real_,
                    nrow = length(lambda_grid),
                    ncol = n_folds,
                    dimnames = list(paste0("lambda_", seq_along(lambda_grid)),
                                    paste0("fold_", seq_len(n_folds))))

  for (f in seq_len(n_folds)) {
    cat("\n=== ", model_name, " | fold ", f, "/", n_folds, " ===\n", sep = "")

    idx_val <- fold_ids[[f]]
    tr_raw  <- data[-idx_val, ]
    va_raw  <- data[idx_val, ]

    # Prep recipe on training fold only
    rec <- make_recipe(tr_raw) %>% prep(training = tr_raw)

    tr_baked <- bake(rec, new_data = NULL)
    va_baked <- bake(rec, new_data = va_raw)

    x_train <- tr_baked %>%
      dplyr::select(-satisfaction) %>%
      as.matrix()

    x_valid <- va_baked %>%
      dplyr::select(-satisfaction) %>%
      as.matrix()

    y_train <- to_binary01(tr_baked$satisfaction)
    y_valid <- va_baked$satisfaction

    for (i in seq_along(lambda_grid)) {
      lam <- lambda_grid[i]

      fit <- glmnet(
        x = x_train,
        y = y_train,
        family = "binomial",
        alpha = alpha_value,
        lambda = lam,
        standardize = FALSE   # already scaled in recipe
      )

      probs <- as.numeric(
        predict(fit, newx = x_valid, type = "response")
      )

      auc_mat[i, f] <- get_auc(probs, y_valid)
    }
  }

  summary_df <- tibble(
    Model = model_name,
    lambda = lambda_grid,
    log_lambda = log(lambda_grid),
    mean_auc = rowMeans(auc_mat, na.rm = TRUE),
    sd_auc = apply(auc_mat, 1, sd, na.rm = TRUE)
  )

  list(
    auc_mat = auc_mat,
    summary = summary_df
  )
}
```
```{r}
lambda_grid <- c(0.0001, 0.001, 0.01, 0.1, 1, 10)
ridge_cv <- cv_glmnet_grid(
  data = cv_data,
  fold_ids = fold_ids,
  make_recipe = make_recipe_full,
  lambda_grid = lambda_grid,
  alpha_value = 0,
  model_name = "Ridge Full"
)

lasso_cv <- cv_glmnet_grid(
  data = cv_data,
  fold_ids = fold_ids,
  make_recipe = make_recipe_full,
  lambda_grid = lambda_grid,
  alpha_value = 1,
  model_name = "Lasso Full"
)
ridge_cv$summary
lasso_cv$summary
```
```{r ridge-cv-plot, fig.height=5, fig.width=8}
ggplot(ridge_cv$summary, aes(x = lambda, y = mean_auc)) +
  geom_ribbon(aes(ymin = mean_auc - sd_auc, ymax = mean_auc + sd_auc),
              alpha = 0.15, fill = "#3498DB") +
  geom_line(color = "#3498DB", linewidth = 1) +
  geom_point(color = "#3498DB", size = 2.5) +
  geom_point(data = ridge_cv$summary %>% slice_max(mean_auc, n = 1, with_ties = FALSE),
             shape = 21, size = 5, stroke = 1.2, color = "#3498DB", fill = "white") +
  geom_text(data = ridge_cv$summary %>% slice_max(mean_auc, n = 1, with_ties = FALSE),
            aes(label = sprintf("%.4f", mean_auc)),
            vjust = -1.4, size = 3.5, color = "#2C3E50") +
  scale_x_log10(labels = scales::label_number()) +
  coord_cartesian(ylim = c(0.80, 0.95)) +
  labs(title = "Ridge (L2) Logistic Regression: 10-Fold CV AUC",
       subtitle = "Shaded band = ± 1 SD; circled point = best λ",
       x = expression(lambda ~ "(log scale)"), y = "Mean CV AUC") +
  theme_minimal(base_size = 12) +
  theme(plot.subtitle = element_text(color = "grey40"))
```

```{r lasso-cv-plot, fig.height=5, fig.width=8}
ggplot(lasso_cv$summary, aes(x = lambda, y = mean_auc)) +
  geom_ribbon(aes(ymin = mean_auc - sd_auc, ymax = mean_auc + sd_auc),
              alpha = 0.15, fill = "#E74C3C") +
  geom_line(color = "#E74C3C", linewidth = 1) +
  geom_point(color = "#E74C3C", size = 2.5) +
  geom_point(data = lasso_cv$summary %>% slice_max(mean_auc, n = 1, with_ties = FALSE),
             shape = 21, size = 5, stroke = 1.2, color = "#E74C3C", fill = "white") +
  geom_text(data = lasso_cv$summary %>% slice_max(mean_auc, n = 1, with_ties = FALSE),
            aes(label = sprintf("%.4f", mean_auc)),
            vjust = -1.4, size = 3.5, color = "#2C3E50") +
  scale_x_log10(labels = scales::label_number()) +
  coord_cartesian(ylim = c(0.45, 0.95)) +
  labs(title = "Lasso (L1) Logistic Regression: 10-Fold CV AUC",
       subtitle = "Shaded band = ± 1 SD; circled point = best λ",
       x = expression(lambda ~ "(log scale)"), y = "Mean CV AUC") +
  theme_minimal(base_size = 12) +
  theme(plot.subtitle = element_text(color = "grey40"))
```

### Why Penalised Methods?

We explored Ridge (L2) and Lasso (L1) regularisation for two reasons. First, penalised methods can improve generalisation by shrinking coefficients toward zero, reducing overfitting — particularly when predictors are correlated (as several service ratings are in this dataset). Second, Lasso has the additional property of performing implicit feature selection by driving some coefficients to exactly zero, offering an alternative to the explicit BIC-based stepwise approach used earlier.

### Results and Interpretation

Both Ridge and Lasso achieve their best AUC at very small $\lambda$ values ($\lambda \leq 0.001$), where the penalty is negligible and the models approximate standard unpenalised logistic regression. At these optimal values, both methods match the unpenalised GLM Full cross-validated AUC of approximately 0.927. As $\lambda$ increases, Ridge degrades smoothly — its AUC declines from 0.927 to 0.836 at $\lambda = 10$ — because L2 regularisation shrinks all coefficients proportionally but never eliminates them entirely. Lasso, by contrast, shows a sharp cliff: at $\lambda = 0.1$ the AUC drops to 0.667, and by $\lambda \geq 1$ it collapses to exactly 0.5 (random guessing) with zero variance across folds, meaning that all predictor coefficients have been driven to zero.

The fact that neither penalised method improves upon the unpenalised logistic regression confirms that the full predictor set is already well-conditioned: there is no meaningful overfitting to regularise away, and the BIC-selected subset (which dropped only 4 weak variables out of 23) was already a near-optimal reduction.

## Threshold Optimisation for Top KNN Models

Cross-validation identified two KNN configurations as the top performers: **KNN (k = 15)** with a mean CV AUC of 0.9777 and mean CV F1 of 0.9118, and **KNN (k = 10)** with a mean CV AUC of 0.9757 and mean CV F1 of 0.9165. Both are very close in performance, and both used the default classification threshold of 0.5. However, the optimal threshold for maximising F1 may differ from 0.5, particularly with moderate class imbalance. We therefore retrain each model on the full training set (`train_with_na`) and sweep across a fine threshold grid on the validation set to identify the threshold that maximises F1 for each model.

```{r threshold-optimisation, fig.height=6, fig.width=9}
# --- Retrain both KNN models on train set, predict on validation set ---
# Preprocess: fit BIC recipe on train_with_na, bake both train and val
rec_bic <- make_recipe_bic(train_with_na) %>% prep(training = train_with_na)
tr_baked_bic <- bake(rec_bic, new_data = NULL)
va_baked_bic <- bake(rec_bic, new_data = val_raw)

tr_x <- tr_baked_bic %>% dplyr::select(-satisfaction) %>% as.data.frame()
va_x <- va_baked_bic %>% dplyr::select(-satisfaction) %>% as.data.frame()
tr_y <- tr_baked_bic$satisfaction
va_y <- va_baked_bic$satisfaction

# KNN k=15
knn15_pred <- knn(tr_x, va_x, cl = tr_y, k = 15, prob = TRUE)
knn15_prob_raw <- attr(knn15_pred, "prob")
prob_knn15 <- ifelse(knn15_pred == "satisfied", knn15_prob_raw, 1 - knn15_prob_raw)

# KNN k=10
knn10_pred <- knn(tr_x, va_x, cl = tr_y, k = 10, prob = TRUE)
knn10_prob_raw <- attr(knn10_pred, "prob")
prob_knn10 <- ifelse(knn10_pred == "satisfied", knn10_prob_raw, 1 - knn10_prob_raw)

# --- Sweep threshold grid ---
threshold_grid <- seq(0.01, 0.99, by = 0.01)

compute_f1_at_threshold <- function(probs, labels, threshold) {
  pred_class <- factor(ifelse(probs >= threshold, "satisfied", "neutral or dissatisfied"),
                       levels = levels(labels))
  cm <- confusionMatrix(pred_class, labels, positive = "satisfied")
  prec <- cm$byClass["Precision"]
  rec  <- cm$byClass["Recall"]
  if (is.na(prec) | is.na(rec) | (prec + rec == 0)) return(0)
  2 * prec * rec / (prec + rec)
}

f1_k15 <- sapply(threshold_grid, function(t) compute_f1_at_threshold(prob_knn15, va_y, t))
f1_k10 <- sapply(threshold_grid, function(t) compute_f1_at_threshold(prob_knn10, va_y, t))

# Build data frame for plotting
f1_df <- bind_rows(
  tibble(Threshold = threshold_grid, F1 = f1_k15, Model = "KNN (k = 15)"),
  tibble(Threshold = threshold_grid, F1 = f1_k10, Model = "KNN (k = 10)")
)

# Find optimal thresholds
best_t15 <- threshold_grid[which.max(f1_k15)]
best_f1_15 <- max(f1_k15)
best_t10 <- threshold_grid[which.max(f1_k10)]
best_f1_10 <- max(f1_k10)

best_points <- tibble(
  Threshold = c(best_t15, best_t10),
  F1 = c(best_f1_15, best_f1_10),
  Model = c("KNN (k = 15)", "KNN (k = 10)"),
  Label = c(sprintf("t* = %.2f, F1 = %.4f", best_t15, best_f1_15),
            sprintf("t* = %.2f, F1 = %.4f", best_t10, best_f1_10))
)

cat("=== Optimal Thresholds (Validation Set) ===\n")
cat(sprintf("KNN k=15: threshold = %.2f, F1 = %.4f\n", best_t15, best_f1_15))
cat(sprintf("KNN k=10: threshold = %.2f, F1 = %.4f\n", best_t10, best_f1_10))

# --- Plot ---
ggplot(f1_df, aes(x = Threshold, y = F1, color = Model)) +
  geom_line(linewidth = 1) +
  geom_point(data = best_points, aes(x = Threshold, y = F1),
             shape = 21, size = 5, stroke = 1.3, fill = "white") +
  geom_text(data = best_points, aes(label = Label),
            vjust = -1.3, hjust = 0.5, size = 3.5, show.legend = FALSE) +
  geom_vline(xintercept = 0.5, linetype = "dashed", color = "grey50", linewidth = 0.5) +
  annotate("text", x = 0.51, y = min(f1_df$F1) + 0.02,
           label = "Default (0.5)", hjust = 0, size = 3, color = "grey40") +
  scale_color_manual(values = c("KNN (k = 15)" = "#3498DB", "KNN (k = 10)" = "#E74C3C")) +
  labs(
    title = "F1 Score vs Classification Threshold — Top 2 KNN Models",
    subtitle = "Trained on full training set; evaluated on validation set; circled points = optimal threshold",
    x = "Classification Threshold", y = "F1 Score",
    color = "Model"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "bottom",
    plot.subtitle = element_text(color = "grey40")
  )
```

# Final Model: KNN (k = 15) — Test-Set Performance

Based on cross-validation and threshold tuning, we select **KNN with k = 15** as the final model. It achieved the highest mean CV AUC (0.9777) and matched the best F1 on the validation set. The optimised threshold from the validation sweep is **t\* = 0.34**. We now evaluate this model on the **held-out test set** (`test_holdout`), which has not been used in any modelling or tuning decision.

```{r final-model-test, fig.height=6, fig.width=8}
# --- Preprocess: fit BIC recipe on train_with_na, bake train and test_holdout ---
rec_final <- make_recipe_bic(train_with_na) %>% prep(training = train_with_na)
tr_baked <- bake(rec_final, new_data = NULL)
te_baked <- bake(rec_final, new_data = test_holdout)

tr_x_final <- tr_baked %>% dplyr::select(-satisfaction) %>% as.data.frame()
te_x_final <- te_baked %>% dplyr::select(-satisfaction) %>% as.data.frame()
tr_y_final <- tr_baked$satisfaction
te_y_final <- te_baked$satisfaction

# --- KNN k=15 prediction on test set ---
set.seed(42)
knn_test_pred <- knn(tr_x_final, te_x_final, cl = tr_y_final, k = 15, prob = TRUE)
knn_test_prob_raw <- attr(knn_test_pred, "prob")
prob_test_satisfied <- ifelse(knn_test_pred == "satisfied",
                              knn_test_prob_raw, 1 - knn_test_prob_raw)

# --- ROC curve on test set ---
roc_test <- roc(te_y_final, prob_test_satisfied, levels = c("neutral or dissatisfied", "satisfied"))

cat("=== Test-Set ROC AUC ===\n")
cat(sprintf("AUC = %.4f\n", auc(roc_test)))

ggroc(roc_test, legacy.axes = TRUE, color = "#3498DB", linewidth = 1.2) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey50") +
  annotate("text", x = 0.55, y = 0.30,
           label = sprintf("AUC = %.4f", auc(roc_test)),
           size = 5, fontface = "bold", color = "#2C3E50") +
  labs(title = "ROC Curve — Final Model: KNN BIC (k = 15) on Held-Out Test Set",
       x = "False Positive Rate (1 − Specificity)",
       y = "True Positive Rate (Sensitivity)") +
  theme_minimal(base_size = 12)
```

## Confusion Matrices and Metrics

We evaluate at both the **default threshold (t = 0.50)** and the **optimised threshold (t = 0.34)** from the validation sweep.

```{r final-confusion-matrices}
# --- Helper: generate confusion matrix and key metrics ---
eval_at_threshold <- function(probs, true_labels, threshold, label) {
  pred_class <- factor(
    ifelse(probs >= threshold, "satisfied", "neutral or dissatisfied"),
    levels = levels(true_labels)
  )
  cm <- confusionMatrix(pred_class, true_labels, positive = "satisfied")

  cat("\n====================================================\n")
  cat(sprintf("  %s (threshold = %.2f)\n", label, threshold))
  cat("====================================================\n")
  print(cm$table)
  cat("\n")
  cat(sprintf("  Accuracy    = %.4f\n", cm$overall["Accuracy"]))
  cat(sprintf("  Sensitivity = %.4f  (Recall)\n", cm$byClass["Sensitivity"]))
  cat(sprintf("  Specificity = %.4f\n", cm$byClass["Specificity"]))
  cat(sprintf("  Precision   = %.4f\n", cm$byClass["Pos Pred Value"]))

  prec <- cm$byClass["Pos Pred Value"]
  rec  <- cm$byClass["Sensitivity"]
  f1   <- ifelse(is.na(prec) | is.na(rec) | (prec + rec == 0), 0,
                 2 * prec * rec / (prec + rec))
  cat(sprintf("  F1 Score    = %.4f\n", f1))

  invisible(list(cm = cm, f1 = f1, pred_class = pred_class))
}

res_050 <- eval_at_threshold(prob_test_satisfied, te_y_final, 0.50, "Test Set — Default")
res_034 <- eval_at_threshold(prob_test_satisfied, te_y_final, 0.34, "Test Set — Optimised")
```

```{r cm-heatmaps, fig.height=4, fig.width=10}
# --- Side-by-side confusion matrix heatmaps ---
make_cm_plot <- function(cm_obj, title_str) {
  cm_df <- as.data.frame(cm_obj$table)
  colnames(cm_df) <- c("Predicted", "Actual", "Freq")
  ggplot(cm_df, aes(x = Actual, y = Predicted, fill = Freq)) +
    geom_tile(color = "white", linewidth = 1.5) +
    geom_text(aes(label = format(Freq, big.mark = ",")), size = 5, fontface = "bold") +
    scale_fill_gradient(low = "#F7F7F7", high = "#3498DB") +
    labs(title = title_str, x = "Actual", y = "Predicted") +
    theme_minimal(base_size = 12) +
    theme(legend.position = "none",
          plot.title = element_text(hjust = 0.5, size = 11))
}

p_cm1 <- make_cm_plot(res_050$cm, "Default Threshold (t = 0.50)")
p_cm2 <- make_cm_plot(res_034$cm, "Optimised Threshold (t = 0.34)")
gridExtra::grid.arrange(p_cm1, p_cm2, ncol = 2,
                        top = grid::textGrob("Confusion Matrices — KNN (k = 15) on Test Set",
                                             gp = grid::gpar(fontsize = 14, fontface = "bold")))
```

## Misclassification Analysis: Inspecting the 15 Nearest Neighbours

To understand *why* the model misclassifies certain observations, we sample 6 misclassified test observations (3 false positives, 3 false negatives under the optimised threshold) and examine the composition and feature statistics of their 15 nearest neighbours in the training set.
```{r}
install.packages("FNN")  # For get.knnx()
```
```{r misclass-analysis}
# --- Identify misclassified observations at t = 0.34 ---
pred_class_034 <- factor(
  ifelse(prob_test_satisfied >= 0.34, "satisfied", "neutral or dissatisfied"),
  levels = levels(te_y_final)
)

misclass_idx <- which(pred_class_034 != te_y_final)
fp_idx <- misclass_idx[te_y_final[misclass_idx] == "neutral or dissatisfied"]
fn_idx <- misclass_idx[te_y_final[misclass_idx] == "satisfied"]

cat(sprintf("Total misclassified at t = 0.34: %d / %d (%.1f%%)\n",
            length(misclass_idx), length(te_y_final),
            100 * length(misclass_idx) / length(te_y_final)))
cat(sprintf("  False positives: %d\n", length(fp_idx)))
cat(sprintf("  False negatives: %d\n", length(fn_idx)))

# --- Sample 3 FP and 3 FN ---
set.seed(123)
fp_sample <- sample(fp_idx, min(3, length(fp_idx)))
fn_sample <- sample(fn_idx, min(3, length(fn_idx)))
sample_idx <- c(fp_sample, fn_sample)

# --- For each sampled observation, find its 15 nearest neighbours in training set ---
# We use Euclidean distance on the baked (preprocessed) feature space
library(FNN)

nn_result <- get.knnx(as.matrix(tr_x_final), as.matrix(te_x_final[sample_idx, ]), k = 15)

# --- Build a summary for each misclassified observation ---
# Map baked column names back for interpretability
baked_colnames <- colnames(tr_x_final)

# For each sampled misclassified point, gather statistics
misclass_report <- list()
for (i in seq_along(sample_idx)) {
  obs_i <- sample_idx[i]
  nn_indices <- nn_result$nn.index[i, ]
  nn_labels  <- tr_y_final[nn_indices]

  # Vote proportions
  n_sat <- sum(nn_labels == "satisfied")
  n_dis <- sum(nn_labels == "neutral or dissatisfied")

  # Features of the test observation (original scale, before baking)
  obs_original <- test_holdout[obs_i, ]

  # Neighbourhood feature statistics (baked scale)
  nn_features <- tr_x_final[nn_indices, ]

  # The test observation features (baked scale)
  obs_baked <- te_x_final[obs_i, ]

  error_type <- ifelse(te_y_final[obs_i] == "neutral or dissatisfied", "FP", "FN")

  misclass_report[[i]] <- list(
    index = obs_i,
    error_type = error_type,
    true_label = as.character(te_y_final[obs_i]),
    predicted_prob = prob_test_satisfied[obs_i],
    nn_satisfied = n_sat,
    nn_dissatisfied = n_dis,
    obs_original = obs_original,
    obs_baked = obs_baked,
    nn_features = nn_features,
    nn_labels = nn_labels
  )
}

# --- Print summary table ---
cat("\n=== Misclassified Observations Summary ===\n\n")
for (r in misclass_report) {
  cat(sprintf("--- Obs #%d | Type: %s | True: %s | P(satisfied) = %.3f ---\n",
              r$index, r$error_type, r$true_label, r$predicted_prob))
  cat(sprintf("    Neighbours: %d satisfied, %d dissatisfied (out of 15)\n",
              r$nn_satisfied, r$nn_dissatisfied))

  # Show key original features
  obs <- r$obs_original
  cat(sprintf("    Class: %s | Travel: %s | Customer: %s | Age: %s\n",
              obs$class, obs$type_of_travel, obs$customer_type, obs$age))
  cat(sprintf("    Flight dist: %s | Dep delay: %s | Arr delay: %s\n",
              obs$flight_distance, obs$departure_delay_in_minutes,
              obs$arrival_delay_in_minutes))

  # Key service ratings
  key_ratings <- c("online_boarding", "inflight_wifi_service",
                   "inflight_entertainment", "seat_comfort",
                   "on_board_service", "leg_room_service")
  rating_vals <- sapply(key_ratings, function(v) as.character(obs[[v]]))
  cat(sprintf("    Key ratings: %s\n",
              paste(paste0(key_ratings, "=", rating_vals), collapse = ", ")))
  cat("\n")
}
```

```{r neighbour-comparison, fig.height=8, fig.width=11}
# --- Compare each misclassified observation to its neighbourhood ---
# For each sampled obs, show how its key features compare to the
# mean of its "satisfied" neighbours vs "dissatisfied" neighbours

key_features_baked <- c("online_boarding", "inflight_wifi_service",
                        "inflight_entertainment", "seat_comfort",
                        "on_board_service", "leg_room_service",
                        "checkin_service", "cleanliness")
# Keep only features that exist in the baked column names
key_features_baked <- intersect(key_features_baked, baked_colnames)

comparison_rows <- list()
for (r in misclass_report) {
  nn_sat <- r$nn_features[r$nn_labels == "satisfied", key_features_baked, drop = FALSE]
  nn_dis <- r$nn_features[r$nn_labels != "satisfied", key_features_baked, drop = FALSE]

  obs_vals <- as.numeric(r$obs_baked[, key_features_baked])

  label_str <- sprintf("Obs #%d (%s)", r$index, r$error_type)

  comparison_rows[[length(comparison_rows) + 1]] <- tibble(
    Observation = label_str,
    Feature = key_features_baked,
    Test_Obs = obs_vals,
    Mean_Sat_NN  = if (nrow(nn_sat) > 0) colMeans(nn_sat) else rep(NA, length(key_features_baked)),
    Mean_Dis_NN  = if (nrow(nn_dis) > 0) colMeans(nn_dis) else rep(NA, length(key_features_baked))
  )
}
comp_df <- bind_rows(comparison_rows)

# Reshape for plotting
comp_long <- comp_df %>%
  pivot_longer(cols = c(Test_Obs, Mean_Sat_NN, Mean_Dis_NN),
               names_to = "Source", values_to = "Value") %>%
  mutate(Source = recode(Source,
    "Test_Obs" = "Misclassified Obs",
    "Mean_Sat_NN" = "Satisfied Neighbours (mean)",
    "Mean_Dis_NN" = "Dissatisfied Neighbours (mean)"
  ))

ggplot(comp_long, aes(x = Feature, y = Value, fill = Source)) +
  geom_col(position = position_dodge(width = 0.75), width = 0.7) +
  facet_wrap(~ Observation, ncol = 2, scales = "free_y") +
  scale_fill_manual(values = c("Misclassified Obs" = "#2C3E50",
                                "Satisfied Neighbours (mean)" = "#3498DB",
                                "Dissatisfied Neighbours (mean)" = "#E74C3C")) +
  labs(
    title = "Misclassified Observations vs Their 15 Nearest Neighbours",
    subtitle = "Key service rating features (baked scale, 0–1); grouped bar = obs value vs neighbour means",
    x = NULL, y = "Feature Value (scaled)",
    fill = NULL
  ) +
  theme_minimal(base_size = 11) +
  theme(
    axis.text.x = element_text(angle = 40, hjust = 1, size = 8),
    legend.position = "bottom",
    strip.text = element_text(face = "bold", size = 10),
    plot.subtitle = element_text(color = "grey40")
  )
```

### Interpretation of Misclassifications

```{r misclass-interpretation}
# --- Quantify how "borderline" the neighbourhood is for each misclassified obs ---
cat("=== Neighbourhood Composition & Margin ===\n\n")
cat(sprintf("%-18s %-5s %-15s %-8s %-8s %-10s\n",
            "Observation", "Type", "True Label", "Sat NN", "Dis NN", "Margin"))
cat(paste(rep("-", 72), collapse = ""), "\n")
for (r in misclass_report) {
  margin <- abs(r$nn_satisfied - r$nn_dissatisfied) / 15
  cat(sprintf("Obs #%-12d %-5s %-15s %-8d %-8d %.3f\n",
              r$index, r$error_type, r$true_label,
              r$nn_satisfied, r$nn_dissatisfied, margin))
}

cat("\nObservations with a low margin (close to 0) sit near the decision boundary,\n")
cat("where the 15 neighbours are nearly evenly split between the two classes.\n")
cat("These are inherently ambiguous cases where the features do not clearly\n")
cat("separate satisfied from dissatisfied passengers.\n")
```

## Final Model Summary

```{r final-summary}
cat("================================================================\n")
cat("  FINAL MODEL SELECTION SUMMARY\n")
cat("================================================================\n\n")
cat("  Model              : KNN (k = 15)\n")
cat("  Data Version       : BIC-reduced (19 predictors from forward stepwise)\n")
cat("  Preprocessing      : Linear imputation, z-score + min-max scaling, dummies\n")
cat(sprintf("  CV AUC (10-fold)   : 0.9777 ± SD\n"))
cat(sprintf("  CV F1 (10-fold)    : 0.9118 ± SD\n"))
cat(sprintf("  Optimised Threshold: 0.34 (from validation set F1 sweep)\n"))
cat(sprintf("  Test AUC           : %.4f\n", auc(roc_test)))
cat(sprintf("  Test F1 (t=0.50)   : %.4f\n", res_050$f1))
cat(sprintf("  Test F1 (t=0.34)   : %.4f\n", res_034$f1))
cat("================================================================\n")
```

KNN with k = 15 on the BIC-reduced predictor set was selected as the final model because it achieved the highest cross-validated AUC (0.9777) among all model–data version configurations and outperformed penalised logistic regression (AUC ≈ 0.927). The model uses the 19 BIC-selected predictors (dummy-encoded), and threshold tuning on the validation set shifted the optimal decision boundary from 0.50 to 0.34, reflecting a slight class imbalance in favour of dissatisfied passengers.

The misclassification analysis reveals that errors tend to occur in **boundary regions** of the feature space where the 15 nearest neighbours contain a near-even mix of both classes. These are passengers whose service ratings and travel characteristics place them between the two clusters — for example, a business traveller who gave high ratings on most service dimensions but low ratings on wifi and online boarding, or a personal traveller with moderate ratings across the board. The low margin (close-to-even neighbour splits) for these observations confirms that they are genuinely ambiguous cases rather than systematic model failures.

The comparison of confusion matrices at t = 0.50 vs t = 0.34 shows that the lower threshold increases sensitivity (catching more truly satisfied passengers) at a modest cost to precision. This trade-off is appropriate if the business priority is to avoid missing satisfied passengers (e.g., for targeted loyalty programs), whereas t = 0.50 would be preferred if false positives are more costly.
