---
title: "Classification Analysis: Airline Passenger Satisfaction"
author: "DSC532 Group Project"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    theme: cosmo
    code_folding: show
    fig_width: 10
    fig_height: 6
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, warning = FALSE, message = FALSE,
  fig.align = "center", fig.width = 10, fig.height = 6
)
```

# Setup and Libraries

```{r libraries}
# Install packages if needed (uncomment on first run)
# install.packages(c("tidyverse", "janitor", "skimr", "corrplot", "caret",
#                     "MASS", "pROC", "knitr", "kableExtra", "gridExtra"))

library(tidyverse)
library(janitor)
library(skimr)
library(corrplot)
library(caret)
library(MASS)
library(pROC)
library(knitr)
library(kableExtra)
library(gridExtra)

set.seed(42)
```

# Data Loading and Overview

```{r load-data}
train_raw <- read_csv("train.csv")
test_raw  <- read_csv("test.csv")

cat("Training set dimensions:", dim(train_raw), "\n")
cat("Test set dimensions:", dim(test_raw), "\n")
```

```{r first-look}
glimpse(train_raw)
```

```{r}
head(train_raw)
```

```{r}
summary(train_raw)
```

# Data Cleaning

## Removing Identifiers and Standardising Column Names

We remove the unnamed index column (`...1`) and the `id` column since they carry no predictive information, and apply `clean_names()` for consistent snake_case naming.

```{r clean-names}
train_clean <- train_raw %>%
  clean_names() %>%
  dplyr::select(-any_of(c("x1", "x_1")), -id)

test_clean <- test_raw %>%
  clean_names() %>%
  dplyr::select(-any_of(c("x1", "x_1")), -id)

colnames(train_clean)
```

## Missing Values Analysis

```{r missing-train}
# Count missing values per variable in the training set
train_clean %>%
  summarise(across(everything(), ~ sum(is.na(.)))) %>%
  pivot_longer(everything(), names_to = "variable", values_to = "na_count") %>%
  arrange(desc(na_count))
```

```{r missing-train-pct}
# Missing percentage per variable in the training set
train_clean %>%
  summarise(across(everything(), ~ mean(is.na(.)))) %>%
  pivot_longer(everything(), names_to = "variable", values_to = "na_percent") %>%
  arrange(desc(na_percent))
```

```{r missing-test}
# Count missing values per variable in the test set
test_clean %>%
  summarise(across(everything(), ~ sum(is.na(.)))) %>%
  pivot_longer(everything(), names_to = "variable", values_to = "na_count") %>%
  arrange(desc(na_count))
```

```{r missing-test-pct}
# Missing percentage per variable in the test set
test_clean %>%
  summarise(across(everything(), ~ mean(is.na(.)))) %>%
  pivot_longer(everything(), names_to = "variable", values_to = "na_percent") %>%
  arrange(desc(na_percent))
```

Missing values are present only in `arrival_delay_in_minutes` — approximately 0.3% of the training set and a similar proportion in the test set. Before deciding on an imputation strategy, we perform a thorough EDA to understand the relationships between `arrival_delay_in_minutes` and all other variables.

# Exploratory Data Analysis

## Target Variable Distribution

```{r target-dist}
train_clean %>%
  count(satisfaction) %>%
  mutate(prop = round(n / sum(n), 3)) %>%
  kable() %>% kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

```{r target-barplot}
ggplot(train_clean, aes(x = satisfaction, fill = satisfaction)) +
  geom_bar(width = 0.6) +
  geom_text(stat = "count", aes(label = after_stat(count)), vjust = -0.5) +
  scale_fill_manual(values = c("neutral or dissatisfied" = "#E74C3C",
                                "satisfied" = "#2ECC71")) +
  labs(title = "Distribution of Passenger Satisfaction",
       x = "Satisfaction", y = "Count") +
  theme_minimal() +
  theme(legend.position = "none")
```

The classes are reasonably balanced (~43% satisfied, ~57% neutral/dissatisfied), so no resampling techniques are strictly necessary.

## Distributions of Continuous Variables

```{r continuous-dist, fig.height=5}
train_clean %>%
  dplyr::select(age, flight_distance,
                departure_delay_in_minutes, arrival_delay_in_minutes) %>%
  pivot_longer(everything()) %>%
  ggplot(aes(x = value)) +
  geom_histogram(bins = 40, fill = "steelblue", color = "white", alpha = 0.8) +
  facet_wrap(~ name, scales = "free", ncol = 2) +
  labs(title = "Distributions of Continuous Variables") +
  theme_minimal()
```

```{r delay-tail-analysis}
# How extreme are the delay tails?
train_clean %>%
  dplyr::select(departure_delay_in_minutes, arrival_delay_in_minutes) %>%
  summarise(across(everything(), list(
    min    = ~ min(., na.rm = TRUE),
    p25    = ~ quantile(., 0.25, na.rm = TRUE),
    median = ~ median(., na.rm = TRUE),
    p75    = ~ quantile(., 0.75, na.rm = TRUE),
    p95    = ~ quantile(., 0.95, na.rm = TRUE),
    p99    = ~ quantile(., 0.99, na.rm = TRUE),
    max    = ~ max(., na.rm = TRUE)
  ))) %>%
  pivot_longer(everything(), names_to = c("variable", "stat"),
               names_sep = "_(?=[^_]+$)") %>%
  pivot_wider(names_from = stat, values_from = value)
```

```{r extreme-counts}
# How many observations are extreme?
train_clean %>%
  summarise(
    dep_over_120  = sum(departure_delay_in_minutes > 120, na.rm = TRUE),
    dep_over_300  = sum(departure_delay_in_minutes > 300, na.rm = TRUE),
    dep_over_500  = sum(departure_delay_in_minutes > 500, na.rm = TRUE),
    arr_over_120  = sum(arrival_delay_in_minutes > 120, na.rm = TRUE),
    arr_over_300  = sum(arrival_delay_in_minutes > 300, na.rm = TRUE),
    arr_over_500  = sum(arrival_delay_in_minutes > 500, na.rm = TRUE)
  )
```

```{r extreme-rows}
# Inspect the actual extreme rows
train_clean %>%
  filter(departure_delay_in_minutes > 500) %>%
  dplyr::select(departure_delay_in_minutes, arrival_delay_in_minutes,
                flight_distance, class, type_of_travel,
                departure_arrival_time_convenient) %>%
  arrange(desc(departure_delay_in_minutes))
```

## Distributions of Service Ratings (Discrete 0–5 Scale)

```{r rating-dist, fig.height=8}
rating_vars <- c("inflight_wifi_service", "departure_arrival_time_convenient",
                  "ease_of_online_booking", "gate_location", "food_and_drink",
                  "online_boarding", "seat_comfort", "inflight_entertainment",
                  "on_board_service", "leg_room_service", "baggage_handling",
                  "checkin_service", "inflight_service", "cleanliness")

train_clean %>%
  dplyr::select(all_of(rating_vars)) %>%
  pivot_longer(everything()) %>%
  ggplot(aes(x = factor(value))) +
  geom_bar(fill = "darkorange", alpha = 0.8) +
  facet_wrap(~ name, scales = "free_y", ncol = 4) +
  labs(title = "Service Rating Distributions (0–5 Scale)", x = "Rating", y = "Count") +
  theme_minimal() +
  theme(strip.text = element_text(size = 8))
```

## Categorical Variable Distributions

```{r cat-dist, fig.height=5}
train_clean %>%
  dplyr::select(gender, customer_type, type_of_travel, class) %>%
  pivot_longer(everything()) %>%
  ggplot(aes(x = value, fill = value)) +
  geom_bar(alpha = 0.8) +
  facet_wrap(~ name, scales = "free", ncol = 2) +
  labs(title = "Categorical Variable Distributions") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1), legend.position = "none")
```

## Satisfaction by Key Categorical Variables

```{r satisfaction-by-cat, fig.height=8}
p1 <- train_clean %>%
  count(class, satisfaction) %>%
  group_by(class) %>%
  mutate(prop = n / sum(n)) %>%
  ggplot(aes(x = class, y = prop, fill = satisfaction)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c("#E74C3C", "#2ECC71")) +
  labs(title = "Satisfaction by Travel Class", y = "Proportion") +
  theme_minimal()

p2 <- train_clean %>%
  count(type_of_travel, satisfaction) %>%
  group_by(type_of_travel) %>%
  mutate(prop = n / sum(n)) %>%
  ggplot(aes(x = type_of_travel, y = prop, fill = satisfaction)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c("#E74C3C", "#2ECC71")) +
  labs(title = "Satisfaction by Travel Type", y = "Proportion") +
  theme_minimal()

p3 <- train_clean %>%
  count(customer_type, satisfaction) %>%
  group_by(customer_type) %>%
  mutate(prop = n / sum(n)) %>%
  ggplot(aes(x = customer_type, y = prop, fill = satisfaction)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c("#E74C3C", "#2ECC71")) +
  labs(title = "Satisfaction by Customer Type", y = "Proportion") +
  theme_minimal()

p4 <- train_clean %>%
  count(gender, satisfaction) %>%
  group_by(gender) %>%
  mutate(prop = n / sum(n)) %>%
  ggplot(aes(x = gender, y = prop, fill = satisfaction)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c("#E74C3C", "#2ECC71")) +
  labs(title = "Satisfaction by Gender", y = "Proportion") +
  theme_minimal()

grid.arrange(p1, p2, p3, p4, ncol = 2)
```

## Age Distribution by Satisfaction

```{r age-dist}
ggplot(train_clean, aes(x = age, fill = satisfaction)) +
  geom_histogram(bins = 40, alpha = 0.6, position = "identity") +
  scale_fill_manual(values = c("#E74C3C", "#2ECC71")) +
  labs(title = "Age Distribution by Satisfaction", x = "Age", y = "Count") +
  theme_minimal()
```

## Boxplots of Service Ratings by Satisfaction

```{r rating-boxplots, fig.height=10}
train_clean %>%
  dplyr::select(all_of(rating_vars), satisfaction) %>%
  pivot_longer(-satisfaction, names_to = "service", values_to = "rating") %>%
  ggplot(aes(x = satisfaction, y = rating, fill = satisfaction)) +
  geom_boxplot(alpha = 0.7) +
  facet_wrap(~ service, ncol = 4) +
  scale_fill_manual(values = c("#E74C3C", "#2ECC71")) +
  labs(title = "Service Ratings by Satisfaction Level") +
  theme_minimal() +
  theme(legend.position = "bottom", strip.text = element_text(size = 8))
```

## Mean Feature Values by Satisfaction Class

```{r mean-by-class}
train_clean %>%
  group_by(satisfaction) %>%
  summarise(across(where(is.numeric), mean, na.rm = TRUE), .groups = "drop") %>%
  pivot_longer(-satisfaction) %>%
  pivot_wider(names_from = satisfaction, values_from = value) %>%
  mutate(diff = satisfied - `neutral or dissatisfied`) %>%
  arrange(desc(abs(diff))) %>%
  kable(digits = 2) %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

## Correlation Analysis

```{r corr-numeric}
# Correlation matrix of all numeric variables
cor_all <- train_clean %>%
  dplyr::select(where(is.numeric)) %>%
  cor(use = "pairwise.complete.obs") %>%
  round(2)

cor_all
```

```{r corr-heatmap-numeric, fig.height=8, fig.width=10}
# Heatmap of numeric correlations
cor_mat_num <- train_clean %>%
  dplyr::select(where(is.numeric)) %>%
  cor(use = "pairwise.complete.obs")

corrplot(cor_mat_num, method = "color", type = "upper", order = "hclust",
         tl.cex = 0.65, tl.col = "black",
         col = colorRampPalette(c("#2166AC", "white", "#B2182B"))(200),
         title = "Correlation Matrix — Numeric Variables",
         mar = c(0, 0, 2, 0))
```

## Arrival Delay: Relationship with All Other Variables

This subsection is key to justifying our imputation strategy. We examine the relationship between `arrival_delay_in_minutes` and every other variable in the dataset.

### Correlation with Numeric Variables

```{r arrival-delay-corr}
# Extract the row of correlations for arrival_delay_in_minutes
arrival_corr <- train_clean %>%
  dplyr::select(where(is.numeric)) %>%
  cor(use = "pairwise.complete.obs") %>%
  as.data.frame() %>%
  rownames_to_column("variable") %>%
  dplyr::select(variable, r = arrival_delay_in_minutes) %>%
  filter(variable != "arrival_delay_in_minutes") %>%
  mutate(abs_r = abs(r)) %>%
  arrange(desc(abs_r))
```

```{r arrival-delay-corr-plot, fig.height=5}
ggplot(arrival_corr, aes(x = reorder(variable, abs_r), y = r)) +
  geom_col(aes(fill = abs_r > 0.1), width = 0.7) +
  scale_fill_manual(values = c("TRUE" = "#E74C3C", "FALSE" = "grey70"),
                    labels = c("TRUE" = "|r| > 0.1", "FALSE" = "|r| ≤ 0.1"),
                    name = "") +
  coord_flip() +
  geom_hline(yintercept = c(-0.1, 0.1), linetype = "dashed", color = "grey40") +
  labs(title = "Correlation of Arrival Delay with All Other Numeric Variables",
       subtitle = "Only departure_delay shows a meaningful correlation (r ≈ 0.97); all others are near zero",
       x = "", y = "Pearson Correlation (r)") +
  theme_minimal()
```

```{r arrival-vs-departure-scatter}
ggplot(train_clean %>% sample_n(5000),
       aes(departure_delay_in_minutes, arrival_delay_in_minutes)) +
  geom_point(alpha = 0.2, color = "steelblue") +
  geom_smooth(method = "lm", color = "red", se = FALSE) +
  labs(title = "Arrival Delay vs Departure Delay",
       subtitle = paste("Pearson r =",
                        round(cor(train_clean$departure_delay_in_minutes,
                                  train_clean$arrival_delay_in_minutes,
                                  use = "complete.obs"), 4))) +
  theme_minimal()
```

The correlation table and bar chart confirm that `departure_delay_in_minutes` is the only variable with a meaningful linear association with `arrival_delay_in_minutes` (r ≈ 0.97). All remaining numeric variables have correlations effectively equal to zero, providing no useful information for imputing missing arrival delays.

### Association with Categorical Variables

We now test whether `arrival_delay_in_minutes` differs across the levels of each categorical variable using both visual (boxplots) and formal statistical tests (Kruskal-Wallis, chosen because the delay distributions are heavily right-skewed and violate the normality assumption of ANOVA).

```{r arrival-delay-boxplots-cat}
train_clean %>%
  dplyr::select(arrival_delay_in_minutes, satisfaction, gender,
                customer_type, type_of_travel, class) %>%
  pivot_longer(-arrival_delay_in_minutes,
               names_to = "variable", values_to = "category") %>%
  ggplot(aes(x = category, y = arrival_delay_in_minutes, fill = category)) +
  geom_boxplot(outlier.alpha = 0.2) +
  facet_wrap(~ variable, scales = "free_x") +
  theme_minimal() +
  theme(legend.position = "none") +
  labs(title = "Arrival Delay by Categorical Variables")
```

```{r arrival-delay-boxplots-zoomed}
# Zoomed-in version (capped at 180 min) to see the bulk of the distribution
train_clean %>%
  dplyr::select(arrival_delay_in_minutes, satisfaction, gender,
                customer_type, type_of_travel, class) %>%
  pivot_longer(-arrival_delay_in_minutes,
               names_to = "variable", values_to = "category") %>%
  ggplot(aes(x = category, y = arrival_delay_in_minutes, fill = category)) +
  geom_boxplot(outlier.alpha = 0.1) +
  facet_wrap(~ variable, scales = "free_x") +
  coord_cartesian(ylim = c(0, 180)) +
  theme_minimal() +
  theme(legend.position = "none") +
  labs(title = "Arrival Delay by Categorical Variables (Zoomed to 0–180 min)",
       y = "Arrival Delay (minutes)", x = "")
```

```{r arrival-delay-mean-by-cat}
# Mean and median arrival delay per categorical group
cat_vars <- c("satisfaction", "gender", "customer_type", "type_of_travel", "class")

for (v in cat_vars) {
  cat("\n── Mean arrival delay by", v, "──\n")
  train_clean %>%
    group_by(.data[[v]]) %>%
    summarise(
      mean_delay = mean(arrival_delay_in_minutes, na.rm = TRUE),
      median_delay = median(arrival_delay_in_minutes, na.rm = TRUE),
      n = n()
    ) %>%
    print()
}
```

### Imputation Justification — Summary

Taken together, this analysis shows that:

1. **Departure delay** is the only variable with a meaningful relationship with arrival delay (r ≈ 0.97, a near-perfect linear association).
2. All other **numeric variables** have correlations effectively equal to zero (|r| < 0.05).
3. The **categorical variables** show no practically meaningful differences in arrival delay across their levels, despite some tests reaching statistical significance due to the large sample size.

Therefore, a simple linear regression model using `departure_delay_in_minutes` as the sole predictor is the most appropriate and parsimonious imputation strategy. Including additional covariates would add no predictive value.

# Missing Value Imputation

We fit a simple linear regression of arrival delay on departure delay using the complete cases in the training data, then use it to impute missing values in both sets. The model is fitted on the training set only to prevent data leakage.

```{r imputation-model}
delay_model <- lm(arrival_delay_in_minutes ~ departure_delay_in_minutes,
                  data = train_clean,
                  na.action = na.exclude)

summary(delay_model)
```

```{r keep-raw-versions}
# Keep versions with missing values for reference
train_with_na <- train_clean
test_with_na  <- test_clean
```

```{r apply-imputation}
# Impute missing values using the fitted model
train_clean$arrival_delay_in_minutes[is.na(train_clean$arrival_delay_in_minutes)] <-
  predict(delay_model,
          newdata = train_clean[is.na(train_clean$arrival_delay_in_minutes), ])

test_clean$arrival_delay_in_minutes[is.na(test_clean$arrival_delay_in_minutes)] <-
  predict(delay_model,
          newdata = test_clean[is.na(test_clean$arrival_delay_in_minutes), ])

cat("Remaining NAs in train:", sum(is.na(train_clean$arrival_delay_in_minutes)), "\n")
cat("Remaining NAs in test:", sum(is.na(test_clean$arrival_delay_in_minutes)), "\n")
```

# Encoding Categorical Variables as Factors

We encode the categorical variables as factors with specified levels. The response variable `satisfaction` is set with "neutral or dissatisfied" as the reference level.

```{r encoding}
encode_factors <- function(df) {
  df %>%
    mutate(
      satisfaction = factor(satisfaction,
                            levels = c("neutral or dissatisfied", "satisfied")),
      gender = factor(gender),
      customer_type = factor(customer_type),
      type_of_travel = factor(type_of_travel),
      class = factor(class, levels = c("Eco", "Eco Plus", "Business"))
    )
}

train <- encode_factors(train_clean)
test  <- encode_factors(test_clean)

str(train[, c("gender", "customer_type", "type_of_travel", "class", "satisfaction")])
```

## Summary of Data Versions

At this point we maintain three versions of the data:

- **`train_raw` / `test_raw`**: original CSV data with all columns and missing values
- **`train_with_na` / `test_with_na`**: cleaned (removed id columns, snake_case names) but still with missing values in `arrival_delay_in_minutes`
- **`train` / `test`**: cleaned, imputed, and factor-encoded — ready for modelling

# Logistic Regression

## Data Preparation

```{r model-prep}
# Verify no NAs remain
cat("Any NAs in train?", any(is.na(train)), "\n")
cat("Any NAs in test?", any(is.na(test)), "\n")

train_y <- train$satisfaction
test_y  <- test$satisfaction
```

## Model Fitting

We fit a logistic regression model with all available predictors. The `glm()` function with `family = binomial` estimates the probability of satisfaction as a function of the predictors via the logit link:

$$\log\left(\frac{P(Y = \text{satisfied} \mid X)}{1 - P(Y = \text{satisfied} \mid X)}\right) = \beta_0 + \beta_1 X_1 + \cdots + \beta_p X_p$$

```{r glm-fit}
glm_fit <- glm(satisfaction ~ ., data = train, family = binomial)
summary(glm_fit)
```
# Interpretation:...

## Predictions and Confusion Matrix

```{r glm-predict}
glm_probs <- predict(glm_fit, newdata = test, type = "response")
glm_pred <- ifelse(glm_probs > 0.5, "satisfied", "neutral or dissatisfied")
glm_pred <- factor(glm_pred, levels = levels(test_y))

glm_cm <- confusionMatrix(glm_pred, test_y, positive = "satisfied")
glm_cm
```

```{r glm-cm-plot, fig.height=5, fig.width=6}
cm_df <- as.data.frame(glm_cm$table)
ggplot(cm_df, aes(x = Reference, y = Prediction, fill = Freq)) +
  geom_tile(color = "white") +
  geom_text(aes(label = Freq), size = 6, fontface = "bold") +
  scale_fill_gradient(low = "white", high = "steelblue") +
  labs(title = "Logistic Regression — Confusion Matrix") +
  theme_minimal() +
  theme(legend.position = "none")
```

## ROC Curve and AUC

```{r glm-roc, fig.height=6}
roc_glm <- roc(test_y, glm_probs,
               levels = c("neutral or dissatisfied", "satisfied"))

plot(roc_glm, col = "#E74C3C", lwd = 2,
     main = "ROC Curve — Logistic Regression")
abline(a = 0, b = 1, lty = 2, col = "grey50")
legend("bottomright",
       legend = paste0("AUC = ", round(auc(roc_glm), 4)),
       col = "#E74C3C", lwd = 2)

cat("AUC:", round(auc(roc_glm), 4), "\n")
```

## Summary Metrics

```{r glm-summary}
tibble(
  Metric = c("Accuracy", "Sensitivity", "Specificity",
             "Balanced Accuracy", "Kappa", "F1 Score", "AUC"),
  Value = c(
    glm_cm$overall["Accuracy"],
    glm_cm$byClass["Sensitivity"],
    glm_cm$byClass["Specificity"],
    glm_cm$byClass["Balanced Accuracy"],
    glm_cm$overall["Kappa"],
    glm_cm$byClass["F1"],
    as.numeric(auc(roc_glm))
  )
) %>%
  kable(digits = 4) %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

```{r session-info}
sessionInfo()
```
