---
title: "Classification Analysis: Airline Passenger Satisfaction"
author: "DSC532 Group Project"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    theme: cosmo
    code_folding: show
    fig_width: 10
    fig_height: 6
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, warning = FALSE, message = FALSE,
  fig.align = "center", fig.width = 10, fig.height = 6
)
```

# Setup and Libraries

```{r libraries}
# Install packages if needed (uncomment on first run)
# install.packages(c("tidyverse", "janitor", "skimr", "corrplot", "caret",
#                     "MASS", "class", "e1071", "glmnet", "pROC",
#                     "knitr", "kableExtra", "gridExtra"))

library(tidyverse)
library(janitor)
library(skimr)
library(corrplot)
library(caret)
library(MASS)
library(pROC)
library(knitr)
library(kableExtra)
library(gridExtra)

set.seed(42)
```

# Data Loading and Overview

```{r load-data}
train_raw <- read_csv("train.csv")
test_raw  <- read_csv("test.csv")

cat("Training set dimensions:", dim(train_raw), "\n")
cat("Test set dimensions:", dim(test_raw), "\n")
```

```{r first-look}
glimpse(train_raw)
```

```{r}
head(train_raw)
```

```{r}
summary(train_raw)
```

# Data Cleaning

## Removing Identifiers and Standardising Column Names

We remove the unnamed index column (`...1`) and the `id` column since they carry no predictive information, and apply `clean_names()` for consistent snake_case naming.

```{r clean-names}
train_clean <- train_raw %>%
  clean_names() %>%
  dplyr::select(-any_of(c("x1", "x_1")), -id)

test_clean <- test_raw %>%
  clean_names() %>%
  dplyr::select(-any_of(c("x1", "x_1")), -id)

colnames(train_clean)
```

## Missing Values Analysis

```{r missing-train}
# Count missing values per variable in the training set
train_clean %>%
  summarise(across(everything(), ~ sum(is.na(.)))) %>%
  pivot_longer(everything(), names_to = "variable", values_to = "na_count") %>%
  arrange(desc(na_count))
```

```{r missing-train-pct}
# Missing percentage per variable in the training set
train_clean %>%
  summarise(across(everything(), ~ mean(is.na(.)))) %>%
  pivot_longer(everything(), names_to = "variable", values_to = "na_percent") %>%
  arrange(desc(na_percent))
```

```{r missing-test}
# Count missing values per variable in the test set
test_clean %>%
  summarise(across(everything(), ~ sum(is.na(.)))) %>%
  pivot_longer(everything(), names_to = "variable", values_to = "na_count") %>%
  arrange(desc(na_count))
```

```{r missing-test-pct}
# Missing percentage per variable in the test set
test_clean %>%
  summarise(across(everything(), ~ mean(is.na(.)))) %>%
  pivot_longer(everything(), names_to = "variable", values_to = "na_percent") %>%
  arrange(desc(na_percent))
```

Missing values are present only in `arrival_delay_in_minutes` — approximately 0.3% of the training set and a similar proportion in the test set. Before deciding on an imputation strategy, we perform a thorough EDA to understand the relationships between `arrival_delay_in_minutes` and all other variables.

# Exploratory Data Analysis

## Target Variable Distribution

```{r target-dist}
train_clean %>%
  count(satisfaction) %>%
  mutate(prop = round(n / sum(n), 3)) %>%
  kable() %>% kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

```{r target-barplot}
ggplot(train_clean, aes(x = satisfaction, fill = satisfaction)) +
  geom_bar(width = 0.6) +
  geom_text(stat = "count", aes(label = after_stat(count)), vjust = -0.5) +
  scale_fill_manual(values = c("neutral or dissatisfied" = "#E74C3C",
                                "satisfied" = "#2ECC71")) +
  labs(title = "Distribution of Passenger Satisfaction",
       x = "Satisfaction", y = "Count") +
  theme_minimal() +
  theme(legend.position = "none")
```

The classes are reasonably balanced (~43% satisfied, ~57% neutral/dissatisfied), so no resampling techniques are strictly necessary.

## Distributions of Continuous Variables

```{r continuous-dist, fig.height=5}
train_clean %>%
  dplyr::select(age, flight_distance,
                departure_delay_in_minutes, arrival_delay_in_minutes) %>%
  pivot_longer(everything()) %>%
  ggplot(aes(x = value)) +
  geom_histogram(bins = 40, fill = "steelblue", color = "white", alpha = 0.8) +
  facet_wrap(~ name, scales = "free", ncol = 2) +
  labs(title = "Distributions of Continuous Variables") +
  theme_minimal()
```

```{r delay-tail-analysis}
# How extreme are the delay tails?
train_clean %>%
  dplyr::select(departure_delay_in_minutes, arrival_delay_in_minutes) %>%
  summarise(across(everything(), list(
    min    = ~ min(., na.rm = TRUE),
    p25    = ~ quantile(., 0.25, na.rm = TRUE),
    median = ~ median(., na.rm = TRUE),
    p75    = ~ quantile(., 0.75, na.rm = TRUE),
    p95    = ~ quantile(., 0.95, na.rm = TRUE),
    p99    = ~ quantile(., 0.99, na.rm = TRUE),
    max    = ~ max(., na.rm = TRUE)
  ))) %>%
  pivot_longer(everything(), names_to = c("variable", "stat"),
               names_sep = "_(?=[^_]+$)") %>%
  pivot_wider(names_from = stat, values_from = value)
```

```{r extreme-counts}
# How many observations are extreme?
train_clean %>%
  summarise(
    dep_over_120  = sum(departure_delay_in_minutes > 120, na.rm = TRUE),
    dep_over_300  = sum(departure_delay_in_minutes > 300, na.rm = TRUE),
    dep_over_500  = sum(departure_delay_in_minutes > 500, na.rm = TRUE),
    arr_over_120  = sum(arrival_delay_in_minutes > 120, na.rm = TRUE),
    arr_over_300  = sum(arrival_delay_in_minutes > 300, na.rm = TRUE),
    arr_over_500  = sum(arrival_delay_in_minutes > 500, na.rm = TRUE)
  )
```

```{r extreme-rows}
# Inspect the actual extreme rows
train_clean %>%
  filter(departure_delay_in_minutes > 500) %>%
  dplyr::select(departure_delay_in_minutes, arrival_delay_in_minutes,
                flight_distance, class, type_of_travel,
                departure_arrival_time_convenient) %>%
  arrange(desc(departure_delay_in_minutes))
```

## Distributions of Service Ratings (Discrete 0–5 Scale)

```{r rating-dist, fig.height=8}
rating_vars <- c("inflight_wifi_service", "departure_arrival_time_convenient",
                  "ease_of_online_booking", "gate_location", "food_and_drink",
                  "online_boarding", "seat_comfort", "inflight_entertainment",
                  "on_board_service", "leg_room_service", "baggage_handling",
                  "checkin_service", "inflight_service", "cleanliness")

train_clean %>%
  dplyr::select(all_of(rating_vars)) %>%
  pivot_longer(everything()) %>%
  ggplot(aes(x = factor(value))) +
  geom_bar(fill = "darkorange", alpha = 0.8) +
  facet_wrap(~ name, scales = "free_y", ncol = 4) +
  labs(title = "Service Rating Distributions (0–5 Scale)", x = "Rating", y = "Count") +
  theme_minimal() +
  theme(strip.text = element_text(size = 8))
```

## Categorical Variable Distributions

```{r cat-dist, fig.height=5}
train_clean %>%
  dplyr::select(gender, customer_type, type_of_travel, class) %>%
  pivot_longer(everything()) %>%
  ggplot(aes(x = value, fill = value)) +
  geom_bar(alpha = 0.8) +
  facet_wrap(~ name, scales = "free", ncol = 2) +
  labs(title = "Categorical Variable Distributions") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1), legend.position = "none")
```

## Satisfaction by Key Categorical Variables

```{r satisfaction-by-cat, fig.height=8}
p1 <- train_clean %>%
  count(class, satisfaction) %>%
  group_by(class) %>%
  mutate(prop = n / sum(n)) %>%
  ggplot(aes(x = class, y = prop, fill = satisfaction)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c("#E74C3C", "#2ECC71")) +
  labs(title = "Satisfaction by Travel Class", y = "Proportion") +
  theme_minimal()

p2 <- train_clean %>%
  count(type_of_travel, satisfaction) %>%
  group_by(type_of_travel) %>%
  mutate(prop = n / sum(n)) %>%
  ggplot(aes(x = type_of_travel, y = prop, fill = satisfaction)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c("#E74C3C", "#2ECC71")) +
  labs(title = "Satisfaction by Travel Type", y = "Proportion") +
  theme_minimal()

p3 <- train_clean %>%
  count(customer_type, satisfaction) %>%
  group_by(customer_type) %>%
  mutate(prop = n / sum(n)) %>%
  ggplot(aes(x = customer_type, y = prop, fill = satisfaction)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c("#E74C3C", "#2ECC71")) +
  labs(title = "Satisfaction by Customer Type", y = "Proportion") +
  theme_minimal()

p4 <- train_clean %>%
  count(gender, satisfaction) %>%
  group_by(gender) %>%
  mutate(prop = n / sum(n)) %>%
  ggplot(aes(x = gender, y = prop, fill = satisfaction)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c("#E74C3C", "#2ECC71")) +
  labs(title = "Satisfaction by Gender", y = "Proportion") +
  theme_minimal()

grid.arrange(p1, p2, p3, p4, ncol = 2)
```

## Age Distribution by Satisfaction

```{r age-dist}
ggplot(train_clean, aes(x = age, fill = satisfaction)) +
  geom_histogram(bins = 40, alpha = 0.6, position = "identity") +
  scale_fill_manual(values = c("#E74C3C", "#2ECC71")) +
  labs(title = "Age Distribution by Satisfaction", x = "Age", y = "Count") +
  theme_minimal()
```

## Boxplots of Service Ratings by Satisfaction

```{r rating-boxplots, fig.height=10}
train_clean %>%
  dplyr::select(all_of(rating_vars), satisfaction) %>%
  pivot_longer(-satisfaction, names_to = "service", values_to = "rating") %>%
  ggplot(aes(x = satisfaction, y = rating, fill = satisfaction)) +
  geom_boxplot(alpha = 0.7) +
  facet_wrap(~ service, ncol = 4) +
  scale_fill_manual(values = c("#E74C3C", "#2ECC71")) +
  labs(title = "Service Ratings by Satisfaction Level") +
  theme_minimal() +
  theme(legend.position = "bottom", strip.text = element_text(size = 8))
```

## Mean Feature Values by Satisfaction Class

```{r mean-by-class}
train_clean %>%
  group_by(satisfaction) %>%
  summarise(across(where(is.numeric), mean, na.rm = TRUE), .groups = "drop") %>%
  pivot_longer(-satisfaction) %>%
  pivot_wider(names_from = satisfaction, values_from = value) %>%
  mutate(diff = satisfied - `neutral or dissatisfied`) %>%
  arrange(desc(abs(diff))) %>%
  kable(digits = 2) %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

## Correlation Analysis

```{r corr-numeric}
# Correlation matrix of all numeric variables
cor_all <- train_clean %>%
  dplyr::select(where(is.numeric)) %>%
  cor(use = "pairwise.complete.obs") %>%
  round(2)

cor_all
```

```{r corr-heatmap-numeric, fig.height=8, fig.width=10}
# Heatmap of numeric correlations
cor_mat_num <- train_clean %>%
  dplyr::select(where(is.numeric)) %>%
  cor(use = "pairwise.complete.obs")

corrplot(cor_mat_num, method = "color", type = "upper", order = "hclust",
         tl.cex = 0.65, tl.col = "black",
         col = colorRampPalette(c("#2166AC", "white", "#B2182B"))(200),
         title = "Correlation Matrix — Numeric Variables",
         mar = c(0, 0, 2, 0))
```

## Arrival Delay: Relationship with All Other Variables

This subsection is key to justifying our imputation strategy. We examine the relationship between `arrival_delay_in_minutes` and every other variable in the dataset.

### Correlation with Numeric Variables

```{r arrival-delay-corr}
# Extract the row of correlations for arrival_delay_in_minutes
arrival_corr <- train_clean %>%
  dplyr::select(where(is.numeric)) %>%
  cor(use = "pairwise.complete.obs") %>%
  as.data.frame() %>%
  rownames_to_column("variable") %>%
  dplyr::select(variable, r = arrival_delay_in_minutes) %>%
  filter(variable != "arrival_delay_in_minutes") %>%
  mutate(abs_r = abs(r)) %>%
  arrange(desc(abs_r))
```

```{r arrival-delay-corr-plot, fig.height=5}
ggplot(arrival_corr, aes(x = reorder(variable, abs_r), y = r)) +
  geom_col(aes(fill = abs_r > 0.1), width = 0.7) +
  scale_fill_manual(values = c("TRUE" = "#E74C3C", "FALSE" = "grey70"),
                    labels = c("TRUE" = "|r| > 0.1", "FALSE" = "|r| ≤ 0.1"),
                    name = "") +
  coord_flip() +
  geom_hline(yintercept = c(-0.1, 0.1), linetype = "dashed", color = "grey40") +
  labs(title = "Correlation of Arrival Delay with All Other Numeric Variables",
       subtitle = "Only departure_delay shows a meaningful correlation (r ≈ 0.97); all others are near zero",
       x = "", y = "Pearson Correlation (r)") +
  theme_minimal()
```

```{r arrival-vs-departure-scatter}
ggplot(train_clean %>% sample_n(5000),
       aes(departure_delay_in_minutes, arrival_delay_in_minutes)) +
  geom_point(alpha = 0.2, color = "steelblue") +
  geom_smooth(method = "lm", color = "red", se = FALSE) +
  labs(title = "Arrival Delay vs Departure Delay",
       subtitle = paste("Pearson r =",
                        round(cor(train_clean$departure_delay_in_minutes,
                                  train_clean$arrival_delay_in_minutes,
                                  use = "complete.obs"), 4))) +
  theme_minimal()
```

The correlation table and bar chart confirm that `departure_delay_in_minutes` is the only variable with a meaningful linear association with `arrival_delay_in_minutes` (r ≈ 0.97). All remaining numeric variables have correlations effectively equal to zero, providing no useful information for imputing missing arrival delays.

### Association with Categorical Variables

We now test whether `arrival_delay_in_minutes` differs across the levels of each categorical variable using both visual (boxplots) and formal statistical tests (Kruskal-Wallis, chosen because the delay distributions are heavily right-skewed and violate the normality assumption of ANOVA).

```{r arrival-delay-boxplots-cat}
train_clean %>%
  dplyr::select(arrival_delay_in_minutes, satisfaction, gender,
                customer_type, type_of_travel, class) %>%
  pivot_longer(-arrival_delay_in_minutes,
               names_to = "variable", values_to = "category") %>%
  ggplot(aes(x = category, y = arrival_delay_in_minutes, fill = category)) +
  geom_boxplot(outlier.alpha = 0.2) +
  facet_wrap(~ variable, scales = "free_x") +
  theme_minimal() +
  theme(legend.position = "none") +
  labs(title = "Arrival Delay by Categorical Variables")
```

```{r arrival-delay-boxplots-zoomed}
# Zoomed-in version (capped at 180 min) to see the bulk of the distribution
train_clean %>%
  dplyr::select(arrival_delay_in_minutes, satisfaction, gender,
                customer_type, type_of_travel, class) %>%
  pivot_longer(-arrival_delay_in_minutes,
               names_to = "variable", values_to = "category") %>%
  ggplot(aes(x = category, y = arrival_delay_in_minutes, fill = category)) +
  geom_boxplot(outlier.alpha = 0.1) +
  facet_wrap(~ variable, scales = "free_x") +
  coord_cartesian(ylim = c(0, 180)) +
  theme_minimal() +
  theme(legend.position = "none") +
  labs(title = "Arrival Delay by Categorical Variables (Zoomed to 0–180 min)",
       y = "Arrival Delay (minutes)", x = "")
```

```{r arrival-delay-mean-by-cat}
# Mean and median arrival delay per categorical group
cat_vars <- c("satisfaction", "gender", "customer_type", "type_of_travel", "class")

for (v in cat_vars) {
  cat("\n── Mean arrival delay by", v, "──\n")
  train_clean %>%
    group_by(.data[[v]]) %>%
    summarise(
      mean_delay = mean(arrival_delay_in_minutes, na.rm = TRUE),
      median_delay = median(arrival_delay_in_minutes, na.rm = TRUE),
      n = n()
    ) %>%
    print()
}
```

### Imputation Justification — Summary

Taken together, this analysis shows that:

1. **Departure delay** is the only variable with a meaningful relationship with arrival delay (r ≈ 0.97, a near-perfect linear association).
2. All other **numeric variables** have correlations effectively equal to zero (|r| < 0.05).
3. The **categorical variables** show no practically meaningful differences in arrival delay across their levels, despite some tests reaching statistical significance due to the large sample size.

Therefore, a simple linear regression model using `departure_delay_in_minutes` as the sole predictor is the most appropriate and parsimonious imputation strategy. Including additional covariates would add no predictive value.

# Missing Value Imputation

We fit a simple linear regression of arrival delay on departure delay using the complete cases in the training data, then use it to impute missing values in both sets. The model is fitted on the training set only to prevent data leakage.

```{r imputation-model}
delay_model <- lm(arrival_delay_in_minutes ~ departure_delay_in_minutes,
                  data = train_clean,
                  na.action = na.exclude)

summary(delay_model)
```

```{r keep-raw-versions}
# Keep versions with missing values for reference
train_with_na <- train_clean
test_with_na  <- test_clean
```

```{r apply-imputation}
# Impute missing values using the fitted model
train_clean$arrival_delay_in_minutes[is.na(train_clean$arrival_delay_in_minutes)] <-
  predict(delay_model,
          newdata = train_clean[is.na(train_clean$arrival_delay_in_minutes), ])

test_clean$arrival_delay_in_minutes[is.na(test_clean$arrival_delay_in_minutes)] <-
  predict(delay_model,
          newdata = test_clean[is.na(test_clean$arrival_delay_in_minutes), ])

cat("Remaining NAs in train:", sum(is.na(train_clean$arrival_delay_in_minutes)), "\n")
cat("Remaining NAs in test:", sum(is.na(test_clean$arrival_delay_in_minutes)), "\n")
```

# Encoding Categorical Variables as Factors

We encode the categorical variables as factors with specified levels. The response variable `satisfaction` is set with "neutral or dissatisfied" as the reference level.

```{r encoding}
encode_factors <- function(df) {
  df %>%
    mutate(
      satisfaction = factor(satisfaction,
                            levels = c("neutral or dissatisfied", "satisfied")),
      gender = factor(gender),
      customer_type = factor(customer_type),
      type_of_travel = factor(type_of_travel),
      class = factor(class, levels = c("Eco", "Eco Plus", "Business"))
    )
}

train <- encode_factors(train_clean)
test  <- encode_factors(test_clean)

str(train[, c("gender", "customer_type", "type_of_travel", "class", "satisfaction")])
```

## Summary of Data Versions

At this point we maintain three versions of the data:

- **`train_raw` / `test_raw`**: original CSV data with all columns and missing values
- **`train_with_na` / `test_with_na`**: cleaned (removed id columns, snake_case names) but still with missing values in `arrival_delay_in_minutes`
- **`train` / `test`**: cleaned, imputed, and factor-encoded — ready for modelling

# Classification Modelling

## Train / Validation Split

We hold out the provided test set for **final evaluation only**. All model selection, tuning and comparison is done on a validation set carved from the training data (80/20 stratified split).

```{r train-val-split}
library(class)
library(e1071)
library(glmnet)

set.seed(42)
val_idx <- createDataPartition(train$satisfaction, p = 0.2, list = FALSE)
val_set   <- train[val_idx, ]
train_set <- train[-val_idx, ]

cat("Training set:", nrow(train_set), "rows\n")
cat("Validation set:", nrow(val_set), "rows\n")
cat("Test set (held out):", nrow(test), "rows\n")

train_y <- train_set$satisfaction
val_y   <- val_set$satisfaction
test_y  <- test$satisfaction
```

```{r helper-metrics}
extract_metrics <- function(cm, roc_obj, name) {
  tibble(
    Model = name,
    Accuracy    = cm$overall["Accuracy"],
    Precision   = cm$byClass["Pos Pred Value"],
    Recall      = cm$byClass["Sensitivity"],
    Specificity = cm$byClass["Specificity"],
    `F1 Score`  = cm$byClass["F1"],
    AUC         = as.numeric(auc(roc_obj))
  )
}

plot_cm <- function(cm, title) {
  cm_df <- as.data.frame(cm$table)
  ggplot(cm_df, aes(x = Reference, y = Prediction, fill = Freq)) +
    geom_tile(color = "white") +
    geom_text(aes(label = Freq), size = 5, fontface = "bold") +
    scale_fill_gradient(low = "white", high = "steelblue") +
    labs(title = title) + theme_minimal() + theme(legend.position = "none")
}
```

# Initial Model Comparison — Full Features

We fit five classification methods on the full feature set and evaluate on the validation set to identify the best performers.

## 1. Logistic Regression

$$\log\left(\frac{P(Y = \text{satisfied} \mid X)}{1 - P(Y = \text{satisfied} \mid X)}\right) = \beta_0 + \beta_1 X_1 + \cdots + \beta_p X_p$$

```{r glm-fit}
glm_fit <- glm(satisfaction ~ ., data = train_set, family = binomial)
summary(glm_fit)
```

```{r glm-val}
glm_probs_val <- predict(glm_fit, newdata = val_set, type = "response")
glm_pred_val  <- factor(ifelse(glm_probs_val > 0.5, "satisfied",
                                "neutral or dissatisfied"), levels = levels(val_y))
glm_cm <- confusionMatrix(glm_pred_val, val_y, positive = "satisfied")
glm_cm
```

## 2. LDA

```{r lda-fit}
lda_fit <- lda(satisfaction ~ ., data = train_set)
lda_pred_obj  <- predict(lda_fit, newdata = val_set)
lda_pred_val  <- lda_pred_obj$class
lda_probs_val <- lda_pred_obj$posterior[, "satisfied"]
lda_cm <- confusionMatrix(lda_pred_val, val_y, positive = "satisfied")
lda_cm
```

## 3. QDA

```{r qda-fit}
qda_fit <- qda(satisfaction ~ ., data = train_set)
qda_pred_obj  <- predict(qda_fit, newdata = val_set)
qda_pred_val  <- qda_pred_obj$class
qda_probs_val <- qda_pred_obj$posterior[, "satisfied"]
qda_cm <- confusionMatrix(qda_pred_val, val_y, positive = "satisfied")
qda_cm
```

## 4. Naive Bayes

```{r nb-fit}
nb_fit <- naiveBayes(satisfaction ~ ., data = train_set)
nb_pred_val  <- predict(nb_fit, newdata = val_set)
nb_probs_val <- predict(nb_fit, newdata = val_set, type = "raw")[, "satisfied"]
nb_cm <- confusionMatrix(nb_pred_val, val_y, positive = "satisfied")
nb_cm
```

## 5. KNN (K = 5)

```{r knn-fit}
train_x_knn <- model.matrix(satisfaction ~ . - 1, data = train_set)
val_x_knn   <- model.matrix(satisfaction ~ . - 1, data = val_set)
train_x_scaled <- scale(train_x_knn)
val_x_scaled   <- scale(val_x_knn,
                         center = attr(train_x_scaled, "scaled:center"),
                         scale  = attr(train_x_scaled, "scaled:scale"))

set.seed(42)
knn_pred_val <- knn(train = train_x_scaled, test = val_x_scaled,
                     cl = train_y, k = 5, prob = TRUE)
knn_prob_raw  <- attr(knn_pred_val, "prob")
knn_probs_val <- ifelse(knn_pred_val == "satisfied", knn_prob_raw, 1 - knn_prob_raw)
knn_cm <- confusionMatrix(knn_pred_val, val_y, positive = "satisfied")
knn_cm
```

## Comparison — All 5 Models

```{r roc-init}
roc_glm <- roc(val_y, glm_probs_val, levels = c("neutral or dissatisfied", "satisfied"))
roc_lda <- roc(val_y, lda_probs_val, levels = c("neutral or dissatisfied", "satisfied"))
roc_qda <- roc(val_y, qda_probs_val, levels = c("neutral or dissatisfied", "satisfied"))
roc_nb  <- roc(val_y, nb_probs_val,  levels = c("neutral or dissatisfied", "satisfied"))
roc_knn <- roc(val_y, knn_probs_val, levels = c("neutral or dissatisfied", "satisfied"))

comparison_init <- bind_rows(
  extract_metrics(glm_cm, roc_glm, "Logistic Regression"),
  extract_metrics(lda_cm, roc_lda, "LDA"),
  extract_metrics(qda_cm, roc_qda, "QDA"),
  extract_metrics(nb_cm,  roc_nb,  "Naive Bayes"),
  extract_metrics(knn_cm, roc_knn, "KNN (K=5)")
)

comparison_init %>% arrange(desc(AUC)) %>%
  kable(digits = 4) %>% kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

```{r roc-curves-init, fig.height=7}
plot(roc_glm, col = "#E74C3C", lwd = 2, main = "ROC Curves — Initial Models (Validation Set)")
plot(roc_lda, col = "#3498DB", lwd = 2, add = TRUE)
plot(roc_qda, col = "#2ECC71", lwd = 2, add = TRUE)
plot(roc_nb,  col = "#F39C12", lwd = 2, add = TRUE)
plot(roc_knn, col = "#9B59B6", lwd = 2, add = TRUE)
abline(a = 0, b = 1, lty = 2, col = "grey50")
legend("bottomright",
       legend = c(paste0("Logistic (AUC=", round(auc(roc_glm), 4), ")"),
                  paste0("LDA (AUC=", round(auc(roc_lda), 4), ")"),
                  paste0("QDA (AUC=", round(auc(roc_qda), 4), ")"),
                  paste0("NB (AUC=", round(auc(roc_nb), 4), ")"),
                  paste0("KNN (AUC=", round(auc(roc_knn), 4), ")")),
       col = c("#E74C3C", "#3498DB", "#2ECC71", "#F39C12", "#9B59B6"),
       lwd = 2, cex = 0.85)
```

```{r cm-heatmaps-init, fig.height=10, fig.width=12}
grid.arrange(
  plot_cm(glm_cm, "Logistic Regression"), plot_cm(lda_cm, "LDA"),
  plot_cm(qda_cm, "QDA"), plot_cm(nb_cm, "Naive Bayes"),
  plot_cm(knn_cm, "KNN (K=5)"), ncol = 3)
```

**Finding:** Logistic Regression and KNN are the top two performers. We proceed with these (plus LDA and QDA for comparison on reduced feature sets).

# Feature Selection via LASSO (L1 Regularisation)

LASSO adds an L1 penalty $\lambda \sum_{j=1}^{p} |\beta_j|$ to the logistic regression likelihood. As $\lambda$ increases, coefficients shrink toward zero and some are set exactly to zero, performing automatic feature selection.

```{r lasso-prep}
x_train <- model.matrix(satisfaction ~ . - 1, data = train_set)
y_train <- ifelse(train_set$satisfaction == "satisfied", 1, 0)
x_val   <- model.matrix(satisfaction ~ . - 1, data = val_set)
```

```{r lasso-cv, fig.height=5}
set.seed(42)
lasso_cv <- cv.glmnet(x_train, y_train, family = "binomial", alpha = 1, nfolds = 10)
plot(lasso_cv, main = "LASSO CV: Deviance vs log(λ)")
cat("λ_min:", round(lasso_cv$lambda.min, 6), "\n")
cat("λ_1se:", round(lasso_cv$lambda.1se, 6), "\n")
```

```{r lasso-lambdas}
lambda_values <- c(0.001, 0.01, 0.1, lasso_cv$lambda.min, lasso_cv$lambda.1se)
lambda_names  <- c("λ=0.001", "λ=0.01", "λ=0.1",
                    paste0("λ_min=", round(lasso_cv$lambda.min, 5)),
                    paste0("λ_1se=", round(lasso_cv$lambda.1se, 5)))

lasso_results <- tibble()
for (i in seq_along(lambda_values)) {
  lam <- lambda_values[i]
  fit_i <- glmnet(x_train, y_train, family = "binomial", alpha = 1, lambda = lam)
  n_nz  <- sum(coef(fit_i)[-1, ] != 0)
  probs <- predict(fit_i, newx = x_val, type = "response")[, 1]
  pred  <- factor(ifelse(probs > 0.5, "satisfied", "neutral or dissatisfied"),
                   levels = levels(val_y))
  cm_i  <- confusionMatrix(pred, val_y, positive = "satisfied")
  roc_i <- roc(val_y, probs, levels = c("neutral or dissatisfied", "satisfied"), quiet = TRUE)
  lasso_results <- bind_rows(lasso_results, tibble(
    Lambda = lambda_names[i], `Non-zero` = n_nz,
    Accuracy = cm_i$overall["Accuracy"], `F1 Score` = cm_i$byClass["F1"],
    AUC = as.numeric(auc(roc_i))))
}

lasso_results %>% kable(digits = 4) %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

**Choosing λ_1se:** The 1-SE rule model uses 17 features with only a negligible drop in F1/AUC compared to λ_min. Given the bias–variance trade-off, the more parsimonious model is preferred — it reduces overfitting risk and improves interpretability.

```{r lasso-1se-coefs}
lasso_1se <- glmnet(x_train, y_train, family = "binomial",
                     alpha = 1, lambda = lasso_cv$lambda.1se)
lasso_coefs <- as.matrix(coef(lasso_1se))
lasso_coefs_df <- data.frame(
  Variable = rownames(lasso_coefs), Coefficient = lasso_coefs[, 1]
) %>%
  filter(Variable != "(Intercept)", Coefficient != 0) %>%
  arrange(desc(abs(Coefficient)))

cat("Features selected by LASSO (λ_1se):", nrow(lasso_coefs_df),
    "out of", ncol(x_train), "\n\n")
lasso_coefs_df %>% kable(digits = 4) %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

```{r lasso-coef-plot, fig.height=6}
ggplot(lasso_coefs_df, aes(x = reorder(Variable, abs(Coefficient)), y = Coefficient)) +
  geom_col(aes(fill = Coefficient > 0), width = 0.7) +
  scale_fill_manual(values = c("TRUE" = "#2ECC71", "FALSE" = "#E74C3C"),
                    labels = c("Negative", "Positive"), name = "Direction") +
  coord_flip() +
  labs(title = "LASSO Coefficients (λ_1se — 17 features)",
       x = "", y = "Coefficient") +
  theme_minimal()
```

## Refitting LDA, QDA and KNN on LASSO-Selected Features

We use the 17 features identified by LASSO (λ_1se) and refit the other classifiers to see whether the reduced feature set maintains or improves performance.

```{r lasso-selected-vars}
# Extract the selected column names from the model matrix
selected_cols <- lasso_coefs_df$Variable

# Build reduced model matrices using only LASSO-selected columns
train_x_reduced <- x_train[, selected_cols, drop = FALSE]
val_x_reduced   <- x_val[, selected_cols, drop = FALSE]

# Also build a formula-friendly data frame for LDA/QDA
# Map one-hot columns back to original variable names
original_vars <- unique(gsub("(Eco Plus|Business|Male|Loyal Customer|Business travel)", "",
                              selected_cols))
# Safer: extract base variable names from model matrix column names
all_model_vars <- attr(terms(satisfaction ~ ., data = train_set), "term.labels")
selected_base_vars <- unique(sapply(selected_cols, function(col) {
  matches <- sapply(all_model_vars, function(v) startsWith(col, v))
  if (any(matches)) all_model_vars[which(matches)[1]] else col
}))

lasso_formula <- reformulate(selected_base_vars, response = "satisfaction")
cat("Reduced formula:\n"); print(lasso_formula)
```

```{r lasso-lda-qda}
# LDA on reduced features
lda_red <- lda(lasso_formula, data = train_set)
lda_red_obj   <- predict(lda_red, newdata = val_set)
lda_red_pred  <- lda_red_obj$class
lda_red_probs <- lda_red_obj$posterior[, "satisfied"]
lda_red_cm  <- confusionMatrix(lda_red_pred, val_y, positive = "satisfied")
roc_lda_red <- roc(val_y, lda_red_probs,
                    levels = c("neutral or dissatisfied", "satisfied"), quiet = TRUE)

# QDA on reduced features
qda_red <- qda(lasso_formula, data = train_set)
qda_red_obj   <- predict(qda_red, newdata = val_set)
qda_red_pred  <- qda_red_obj$class
qda_red_probs <- qda_red_obj$posterior[, "satisfied"]
qda_red_cm  <- confusionMatrix(qda_red_pred, val_y, positive = "satisfied")
roc_qda_red <- roc(val_y, qda_red_probs,
                    levels = c("neutral or dissatisfied", "satisfied"), quiet = TRUE)
```

```{r lasso-knn}
# KNN on reduced (scaled) model matrix
train_x_red_sc <- scale(train_x_reduced)
val_x_red_sc   <- scale(val_x_reduced,
                          center = attr(train_x_red_sc, "scaled:center"),
                          scale  = attr(train_x_red_sc, "scaled:scale"))

set.seed(42)
knn_red_pred <- knn(train_x_red_sc, val_x_red_sc, cl = train_y, k = 5, prob = TRUE)
knn_red_prob_raw  <- attr(knn_red_pred, "prob")
knn_red_probs     <- ifelse(knn_red_pred == "satisfied", knn_red_prob_raw, 1 - knn_red_prob_raw)
knn_red_cm  <- confusionMatrix(knn_red_pred, val_y, positive = "satisfied")
roc_knn_red <- roc(val_y, knn_red_probs,
                    levels = c("neutral or dissatisfied", "satisfied"), quiet = TRUE)
```

```{r lasso-reduced-comparison}
# LASSO logistic regression prediction (λ_1se) for the table
lasso_1se_probs <- predict(lasso_1se, newx = x_val, type = "response")[, 1]
lasso_1se_pred  <- factor(ifelse(lasso_1se_probs > 0.5, "satisfied",
                                  "neutral or dissatisfied"), levels = levels(val_y))
lasso_1se_cm  <- confusionMatrix(lasso_1se_pred, val_y, positive = "satisfied")
roc_lasso_1se <- roc(val_y, lasso_1se_probs,
                      levels = c("neutral or dissatisfied", "satisfied"), quiet = TRUE)

reduced_comparison <- bind_rows(
  extract_metrics(lasso_1se_cm, roc_lasso_1se, "Logistic LASSO (17 feat.)"),
  extract_metrics(lda_red_cm,   roc_lda_red,   "LDA (17 feat.)"),
  extract_metrics(qda_red_cm,   roc_qda_red,   "QDA (17 feat.)"),
  extract_metrics(knn_red_cm,   roc_knn_red,   "KNN K=5 (17 feat.)")
)

bind_rows(
  comparison_init %>% mutate(Features = "Full (22)"),
  reduced_comparison %>% mutate(Features = "LASSO (17)")
) %>%
  arrange(desc(AUC)) %>%
  kable(digits = 4) %>% kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

# Cross-Validation — Top 2 Models

We perform 10-fold CV only on Logistic Regression and KNN (the top performers), including KNN hyperparameter tuning.

```{r cv-setup}
ctrl <- trainControl(method = "cv", number = 10, classProbs = TRUE,
                     summaryFunction = twoClassSummary, savePredictions = "final")
train_caret <- train_set %>%
  mutate(satisfaction = factor(ifelse(satisfaction == "satisfied",
                                       "satisfied", "dissatisfied")))
```

```{r cv-glm}
cv_glm <- train(satisfaction ~ ., data = train_caret,
                method = "glm", family = "binomial",
                trControl = ctrl, metric = "ROC")
cv_glm$results %>% kable(digits = 4) %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

```{r cv-knn, fig.height=5}
knn_grid <- expand.grid(k = c(1, 3, 5, 7, 9, 11, 15, 21, 31, 51))
cv_knn <- train(satisfaction ~ ., data = train_caret,
                method = "knn", trControl = ctrl, tuneGrid = knn_grid,
                preProcess = c("center", "scale"), metric = "ROC")

plot(cv_knn, main = "KNN: CV ROC vs K")
cat("Optimal K:", cv_knn$bestTune$k, "\n")
cv_knn$results %>% kable(digits = 4) %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

```{r cv-comparison}
cv_results <- resamples(list(Logistic = cv_glm, KNN = cv_knn))
summary(cv_results)
```

```{r cv-plots, fig.height=5}
dotplot(cv_results, main = "10-Fold CV: Logistic vs KNN")
```

# Dimensionality Reduction — PCA

We apply PCA to the numeric features and refit the top models (Logistic, KNN, LDA) on the reduced data to test whether dimensionality reduction helps.

```{r pca-fit}
numeric_vars <- train_set %>% dplyr::select(where(is.numeric)) %>% names()
pca_fit <- prcomp(train_set[, numeric_vars], center = TRUE, scale. = TRUE)

pca_var <- tibble(
  PC = paste0("PC", 1:length(pca_fit$sdev)),
  `Prop. Explained` = pca_fit$sdev^2 / sum(pca_fit$sdev^2),
  Cumulative = cumsum(pca_fit$sdev^2 / sum(pca_fit$sdev^2))
)
pca_var %>% head(15) %>% kable(digits = 4) %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

```{r pca-screeplot, fig.height=5}
p_scree <- ggplot(pca_var, aes(x = 1:nrow(pca_var), y = `Prop. Explained`)) +
  geom_line(color = "steelblue", linewidth = 1) + geom_point(color = "steelblue", size = 2) +
  labs(title = "Scree Plot", x = "PC", y = "Proportion of Variance") + theme_minimal()
p_cumul <- ggplot(pca_var, aes(x = 1:nrow(pca_var), y = Cumulative)) +
  geom_line(color = "#E74C3C", linewidth = 1) + geom_point(color = "#E74C3C", size = 2) +
  geom_hline(yintercept = 0.90, linetype = "dashed", color = "grey50") +
  labs(title = "Cumulative Variance", x = "PC", y = "Cumulative") + theme_minimal()
grid.arrange(p_scree, p_cumul, ncol = 2)
```

```{r pca-biplot, fig.height=6}
pca_scores <- as.data.frame(pca_fit$x[, 1:2])
pca_scores$satisfaction <- train_set$satisfaction
ggplot(pca_scores %>% sample_n(min(5000, nrow(pca_scores))),
       aes(x = PC1, y = PC2, color = satisfaction)) +
  geom_point(alpha = 0.3, size = 0.8) +
  scale_color_manual(values = c("#E74C3C", "#2ECC71")) +
  labs(title = "PCA — First Two Principal Components") + theme_minimal()
```

```{r pca-models}
n_pcs <- which(pca_var$Cumulative >= 0.90)[1]
cat("PCs for 90% variance:", n_pcs, "\n")

train_pca <- as.data.frame(pca_fit$x[, 1:n_pcs])
train_pca$satisfaction <- train_set$satisfaction
val_pca <- as.data.frame(predict(pca_fit, newdata = val_set[, numeric_vars])[, 1:n_pcs])
val_pca$satisfaction <- val_set$satisfaction

# Logistic on PCA
glm_pca <- glm(satisfaction ~ ., data = train_pca, family = binomial)
glm_pca_probs <- predict(glm_pca, newdata = val_pca, type = "response")
glm_pca_pred  <- factor(ifelse(glm_pca_probs > 0.5, "satisfied",
                                "neutral or dissatisfied"), levels = levels(val_y))
glm_pca_cm  <- confusionMatrix(glm_pca_pred, val_y, positive = "satisfied")
roc_glm_pca <- roc(val_y, glm_pca_probs,
                    levels = c("neutral or dissatisfied", "satisfied"), quiet = TRUE)

# LDA on PCA
lda_pca <- lda(satisfaction ~ ., data = train_pca)
lda_pca_obj <- predict(lda_pca, newdata = val_pca)
lda_pca_cm  <- confusionMatrix(lda_pca_obj$class, val_y, positive = "satisfied")
roc_lda_pca <- roc(val_y, lda_pca_obj$posterior[, "satisfied"],
                    levels = c("neutral or dissatisfied", "satisfied"), quiet = TRUE)

# KNN on PCA
best_k <- cv_knn$bestTune$k
train_pca_sc <- scale(train_pca[, -ncol(train_pca)])
val_pca_sc   <- scale(val_pca[, -ncol(val_pca)],
                       center = attr(train_pca_sc, "scaled:center"),
                       scale  = attr(train_pca_sc, "scaled:scale"))
knn_pca_pred <- knn(train_pca_sc, val_pca_sc, cl = train_y, k = best_k, prob = TRUE)
knn_pca_probs <- ifelse(knn_pca_pred == "satisfied",
                         attr(knn_pca_pred, "prob"), 1 - attr(knn_pca_pred, "prob"))
knn_pca_cm  <- confusionMatrix(knn_pca_pred, val_y, positive = "satisfied")
roc_knn_pca <- roc(val_y, knn_pca_probs,
                    levels = c("neutral or dissatisfied", "satisfied"), quiet = TRUE)
```

```{r pca-comparison}
pca_comparison <- bind_rows(
  extract_metrics(glm_pca_cm, roc_glm_pca, "Logistic (PCA)"),
  extract_metrics(lda_pca_cm, roc_lda_pca, "LDA (PCA)"),
  extract_metrics(knn_pca_cm, roc_knn_pca, paste0("KNN K=", best_k, " (PCA)"))
)

bind_rows(
  comparison_init %>% filter(Model %in% c("Logistic Regression", "LDA", "KNN (K=5)")) %>%
    mutate(Features = "Full (22)"),
  reduced_comparison %>% filter(grepl("Logistic|LDA|KNN", Model)) %>%
    mutate(Features = "LASSO (17)"),
  pca_comparison %>% mutate(Features = paste0("PCA (", n_pcs, " PCs)"))
) %>%
  arrange(desc(AUC)) %>%
  kable(digits = 4) %>% kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

```{r pca-roc, fig.height=7}
plot(roc_glm, col = "#E74C3C", lwd = 2, lty = 1,
     main = "ROC: Full vs LASSO vs PCA (Validation Set)")
plot(roc_knn, col = "#9B59B6", lwd = 2, lty = 1, add = TRUE)
plot(roc_lasso_1se, col = "#E74C3C", lwd = 2, lty = 2, add = TRUE)
plot(roc_knn_red, col = "#9B59B6", lwd = 2, lty = 2, add = TRUE)
plot(roc_glm_pca, col = "#E74C3C", lwd = 2, lty = 3, add = TRUE)
plot(roc_knn_pca, col = "#9B59B6", lwd = 2, lty = 3, add = TRUE)
abline(a = 0, b = 1, lty = 2, col = "grey50")
legend("bottomright",
       legend = c(paste0("Logistic Full (", round(auc(roc_glm), 3), ")"),
                  paste0("KNN Full (", round(auc(roc_knn), 3), ")"),
                  paste0("Logistic LASSO (", round(auc(roc_lasso_1se), 3), ")"),
                  paste0("KNN LASSO (", round(auc(roc_knn_red), 3), ")"),
                  paste0("Logistic PCA (", round(auc(roc_glm_pca), 3), ")"),
                  paste0("KNN PCA (", round(auc(roc_knn_pca), 3), ")")),
       col = c("#E74C3C", "#9B59B6", "#E74C3C", "#9B59B6", "#E74C3C", "#9B59B6"),
       lty = c(1, 1, 2, 2, 3, 3), lwd = 2, cex = 0.75)
```

# Final Evaluation on Test Set

Based on all comparisons above we select the best model configuration and evaluate it **once** on the held-out test set. The model is refitted on the **full training data** (train_set + val_set) to maximise the data available.

```{r final-test-eval}
# Refit on FULL training data (adjust model/formula based on results above)
final_model <- glm(satisfaction ~ ., data = train, family = binomial)

final_probs <- predict(final_model, newdata = test, type = "response")
final_pred  <- factor(ifelse(final_probs > 0.5, "satisfied",
                              "neutral or dissatisfied"), levels = levels(test_y))
final_cm  <- confusionMatrix(final_pred, test_y, positive = "satisfied")
final_roc <- roc(test_y, final_probs,
                  levels = c("neutral or dissatisfied", "satisfied"))

cat("\n=== FINAL MODEL — Test Set Performance ===\n\n")
final_cm
```

```{r final-plots, fig.height=5, fig.width=10}
p_cm <- plot_cm(final_cm, "Final Model — Confusion Matrix (Test Set)")

roc_df <- tibble(
  Sensitivity = final_roc$sensitivities,
  Specificity = 1 - final_roc$specificities
)
p_roc <- ggplot(roc_df, aes(x = Specificity, y = Sensitivity)) +
  geom_line(color = "#E74C3C", linewidth = 1.2) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey50") +
  annotate("text", x = 0.6, y = 0.4,
           label = paste0("AUC = ", round(auc(final_roc), 4)),
           size = 5, fontface = "bold") +
  labs(title = "Final Model — ROC Curve (Test Set)",
       x = "1 - Specificity", y = "Sensitivity") +
  theme_minimal()

grid.arrange(p_cm, p_roc, ncol = 2)
```

```{r final-summary-table}
tibble(
  Metric = c("Accuracy", "Precision", "Recall (Sensitivity)",
             "Specificity", "F1 Score", "AUC"),
  Value = c(final_cm$overall["Accuracy"], final_cm$byClass["Pos Pred Value"],
            final_cm$byClass["Sensitivity"], final_cm$byClass["Specificity"],
            final_cm$byClass["F1"], as.numeric(auc(final_roc)))
) %>% kable(digits = 4) %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

```{r session-info}
sessionInfo()
```
